---
title: "Testing Phylogenetic Signal in Phenotypic Data"
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggtree)
library(phylosignal)
library(broom)
library(rstatix)
library(adephylo)
library(ape)
library(phylobase)
library(formattable)
library(corrplot)
library(ggbeeswarm)
library(cowplot)
library(ggpubr)
library(grid)
library(gridGraphics)
library(ggplot2)
library(readr) 
library(dplyr)
library(tidyr)
library(patchwork)
library(phytools)
library(gridExtra)  
library(knitr)
`%!in%` <- negate(`%in%`)
```

## Prepare data for analysis 
Import phenotype and phylogenetic tree data
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Import phylogenetic tree 
tr <- read.tree("/Users/spencertoth/Desktop/Research/gardnerella_phylogenetic_signal/rootedGardPhylogeny_noOutgroup.tre")
is.rooted(tr)
# Rename tree nodes to match strain names in strain-phenotype data 
numeric_tips <- tr$tip.label[tr$tip.label %in% c("101", "5-1")]
tr$tip.label[match(numeric_tips, tr$tip.label)] <- paste0("Gv", numeric_tips)
#tr$tip.label[tr$tip.label == "14018"] <- "S14018"
tip_labels <- tr$tip.label
is.rooted(tr)

# Append genomospecies to strain name 
strainIDs <- read.csv("/Users/spencertoth/Desktop/Research/gardnerella_phylogenetic_signal/gardnerella_pathogenic_phenotypes/strain_phylogeny/strainIDs.csv",
                      stringsAsFactors = FALSE)
strainIDs$strain <- as.character(strainIDs$strain)
tip_labels <- tr$tip.label
idx <- match(tip_labels, strainIDs$strain)
new_tip_labels <- paste0(tip_labels, "_", strainIDs$genomospecies[idx])
tr$tip.label <- new_tip_labels
```

Average the phenotype values across technical replicates per strain  
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Phenotype file paths
growth_rate_data_file <- read_csv("../formatted_phenotype_data/growth_rate.csv")
car_cap_data_file <- read_csv("../formatted_phenotype_data/carrying_capacity.csv")
percent_biofilm_od_data_file  <- read_csv("../formatted_phenotype_data/percent_biofilm_od.csv")
safranin_od_data_file <- read_csv("../formatted_phenotype_data/safranin_od.csv")
mic_data_file <-read.csv("../formatted_phenotype_data/metro_mic.csv")
pH_growth_data_file <- read_csv("../formatted_phenotype_data/pH_growth.csv")

add_gs <- function(df) {
  df %>%
    mutate(Strain = as.character(Strain)) %>%
    left_join(strainIDs %>%
                select(strain, genomospecies),
              by = c("Strain" = "strain")) %>%
    mutate(Strain = paste0(Strain, "_", genomospecies)) %>%
    select(-genomospecies)
}

growth_rate_data <- growth_rate_data_file %>%
  add_gs() %>%
  #mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_mSlope = mean(mSlope, na.rm = TRUE))
gr_vector <- setNames(growth_rate_data$mean_mSlope, growth_rate_data$Strain)

car_cap_data <- car_cap_data_file %>%
 # mutate(Strain = as.character(Strain)) %>%
  add_gs() %>%
  group_by(Strain) %>% summarise(mean_maxOD = mean(maxOD, na.rm = TRUE))
cc_vector <- setNames(car_cap_data$mean_maxOD, car_cap_data$Strain)

percent_biofilm_od_data <- percent_biofilm_od_data_file %>%
   add_gs() %>%
   mutate(arcsq_mPercentOD = asin(sqrt(mPercentOD/100))) %>%
  group_by(Strain) %>% 
  summarise(mean_mPercentOD = mean(arcsq_mPercentOD, na.rm = TRUE))   
pb_vector <- setNames(percent_biofilm_od_data$mean_mPercentOD, percent_biofilm_od_data$Strain)

safranin_od_data <- safranin_od_data_file %>%
  #mutate(Strain = as.character(Strain)) %>%
  add_gs() %>%
  group_by(Strain) %>% summarise(mean_mSafOD = mean(mSafOD, na.rm = TRUE))
sf_vector <- setNames(safranin_od_data$mean_mSafOD, safranin_od_data$Strain)


#Filter out illegible MIC strips
metro_mic_filtered <- mic_data_file %>%
  filter(!grepl("*to read",
                MIC, ignore.case = TRUE))

# Binarize the filtered MICs (MIC >=32 = 1, <32 = 0)
mic_binarized <- metro_mic_filtered %>%
  mutate(
    Strain = as.character(Strain),
    MIC_binary = case_when(
      grepl("^>256$", MIC) ~ 1,
      grepl("^[0-9]+$", MIC) & as.numeric(MIC) >= 32 ~ 1,
      TRUE ~ 0
    )
  )

# Average binary MIC per strain  
mic_data <- mic_binarized %>%
  add_gs() %>%
   group_by(Strain) %>%
  summarise(
    mean_MIC_binary = mean(MIC_binary, na.rm = TRUE),
    .groups = "drop"
  )
mic_vector <- setNames(mic_data$mean_MIC_binary, mic_data$Strain)



pH_growth_data <- pH_growth_data_file %>%
  group_by(Strain, pH) %>%
  add_gs() %>%
  t_test(mOD ~ 1, mu = 0.01, alternative = "greater") %>%
  add_significance("p") %>%
  filter(p.signif != "ns") %>%
  group_by(Strain) %>%
  filter(pH == min(pH)) %>%
  mutate(pH = as.numeric(pH))
ph_vector <- setNames(pH_growth_data$pH, pH_growth_data$Strain)

```
## Update: "phylosignal" R-package used to conduct phylogenetic signal tests (Moran's I and Pagel’s lambda) 
Subset phylogenetic tree to the strains with phenotype information 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Phenotypes with all 22 strains: carrying capacity, growth rate and minimum pH
data_tips <- intersect(names(gr_vector), tr$tip.label)  
tr_subset  <- keep.tip(tr, data_tips)

tip_labels <- tr_subset$tip.label
n_strains <- length(tip_labels) # 22

p <- ggtree(tr_subset) +
  geom_rootedge(rootedge = 0.05, colour = "black", linewidth = 1) +
  geom_tiplab(size = 4) +
  xlim(NA, 0.75)
p

# Phenotypes with 21 strains: safranin and % biofilm 
data_tips_02 <- intersect(names(sf_vector), tr$tip.label)  
tr_subset_02  <- keep.tip(tr, data_tips_02)
tip_labels_02 <- tr_subset_02$tip.label
n_strains_02 <- length(tip_labels_02) # 21
```

```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
tr_subset_ultra <- force.ultrametric(tr_subset, method = "nnls") # nnls or extend
tr_subset_ultra$edge.length[tr_subset_ultra$edge.length <= 0] <- 1e-6


p <- ggtree(tr_subset_ultra) +
  geom_rootedge(rootedge = 0.05, colour = "black", linewidth = 1) +
  geom_tiplab(size = 4) +
  xlim(NA, 0.75)
p

tr_subset_ultra_02 <- force.ultrametric(tr_subset_02, method = "nnls") # nnls or extend
tr_subset_ultra_02$edge.length[tr_subset_ultra_02$edge.length <= 0] <- 1e-6
``` 

Moran's I and Pagel's Lambda: Phylogenetic Signal along BM Gradient 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Distribution of Test Statistic as percentage of BM increases
set.seed(42)
phylosim_ultra <- phyloSim(tree = tr_subset_ultra, methods = c("I", "Lambda"), model = "BM", pb = FALSE)
phylosim <- phyloSim(tree = tr_subset, methods = c("I", "Lambda"), model = "BM", pb = FALSE)
```

Phylogenetic Signal Statistics Along BM-gradient (Ultrametric vs. Additive Trees)
```{r, message=FALSE, warning=FALSE, echo=TRUE}
plot(phylosim_ultra, stacked.methods = FALSE, quantiles = c(0, 1), ylab = "Phylogenetic Signal Value")
plot.phylosim(phylosim_ultra, what = "pval", stacked.methods = TRUE)
mean(phylosim_ultra$sim.stat[1, 1, ])  # I at 0% BM 
mean(phylosim_ultra$sim.stat[1, 2, ])   # Lambda at 0% BM 

# 100% BM = LAST BM level = column 101
mean_100bm_I <- mean(phylosim_ultra$sim.stat[101, 1, ])      # Moran's I at 100% BM
mean_100bm_lambda <- mean(phylosim_ultra$sim.stat[101, 2, ]) # Pagel's λ at 100% BM

print(paste("Moran's I at 100% BM:", round(mean_100bm_I, 4)))
print(paste("Pagel's λ at 100% BM:", round(mean_100bm_lambda, 4)))

plot(phylosim, stacked.methods = FALSE, quantiles = c(0, 1), ylab = "Phylogenetic Signal Value")
plot.phylosim(phylosim, what = "pval", stacked.methods = TRUE)

# 0% BM (random)
mean(phylosim$sim.stat[1, 1, ])   # I at 0% BM 
mean(phylosim$sim.stat[1, 2, ])   # Lambda at 0% BM 

# 100% BM (last level)
mean_100bm_I <- mean(phylosim$sim.stat[101, 1, ])      # Moran's I
mean_100bm_lambda <- mean(phylosim$sim.stat[101, 2, ]) # Pagel's λ

print(paste("Moran's I at 100% BM:", round(mean_100bm_I, 4)))
print(paste("Pagel's λ at 100% BM:", round(mean_100bm_lambda, 4)))

``` 

## Exponential Growth Rate
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
# Permute the actual growth_rate values (scramble across tree tips)
random <- sample(gr_vector, size = length(tip_labels), replace = FALSE)

# Scale to match observed growth_rate moments
observed <- gr_vector[match(tip_labels, names(gr_vector))]
target_mean <- mean(observed)
target_sd <- sd(observed)



# Brownian Motion for Moran's I Test (Non-Ultrametric)
bm_raw <- rTraitCont(tr_subset, model = "BM")
bm <- (bm_raw - mean(bm_raw)) / sd(bm_raw) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm), sd(bm))

dat_growth_rate <- data.frame(
  growth_rate = gr_vector[match(tip_labels, names(gr_vector))],   
  random      = random,
  bm          = bm
  
)
rownames(dat_growth_rate) <- tip_labels
p4d_growth_rate <- phylo4d(tr_subset, dat_growth_rate)


# Brownian Motion Pagel's Lambda (Ultrametric Tree)
bm_raw_ultra <- rTraitCont(tr_subset_ultra, model = "BM")
bm_ultra <- (bm_raw_ultra - mean(bm_raw_ultra)) / sd(bm_raw_ultra) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm_ultra), sd(bm_ultra))

dat_growth_rate_ultra <- data.frame(
  growth_rate = gr_vector[match(tip_labels, names(gr_vector))],   
  random      = random,
  bm          = bm_ultra
  
)
rownames(dat_growth_rate_ultra) <- tip_labels

p4d_growth_rate_ultra <- phylo4d(tr_subset_ultra, dat_growth_rate_ultra)
# Trait Distribution
barplot.phylo4d(p4d_growth_rate_ultra, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_growth_rate_ultra, tree.type = "phylo", tree.ladderize = TRUE)
```

Moran's I
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_growth_rate, methods = c("I"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
growth_rate.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "growth_rate")
random.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(growth_rate.crlg, main = "Phenotypic Trait: Growth Rate")


#Local Indicator of Phylogenetic Association (LIPA)
growth_rate.lipa <- lipaMoran(p4d_growth_rate)
growth_rate.lipa.p4d <- lipaMoran(p4d_growth_rate, as.p4d = TRUE)

barplot.phylo4d(p4d_growth_rate, bar.col = (growth_rate.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)
barplot.phylo4d(growth_rate.lipa.p4d, bar.col = (growth_rate.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)


lipa_df <- as.data.frame(growth_rate.lipa)
print(lipa_df)
# Filter significant p-values (≤ 0.05)
sig_lipa <- lipa_df[lipa_df$p.value.growth_rate <= 0.05, ]
# Print table
print(sig_lipa)
kable(sig_lipa, caption = "Significant LIPA Results (p ≤ 0.05)")


lipa_df$q.value.growth_rate <- p.adjust(lipa_df$p.value.growth_rate, "BH")
lipa_df$q.value.random <- p.adjust(lipa_df$p.value.random, "BH")
lipa_df$q.value.bm    <- p.adjust(lipa_df$p.value.bm, "BH")


sig_lipa <- lipa_df[
  lipa_df$q.value.growth_rate <= 0.05 | 
  lipa_df$q.value.random <= 0.05 | 
  lipa_df$q.value.bm <= 0.05, 
]
# Print table
print(sig_lipa)
kable(sig_lipa, caption = "FDR adjusted p-value: Significant LIPA Results (q ≤ 0.05)")


tip_order <- match(growth_rate.lipa.p4d@label, rownames(lipa_df))
col_growth <- (lipa_df[tip_order, "q.value.growth_rate"] < 0.05) + 1
col_random <- (lipa_df[tip_order, "q.value.random"] < 0.05) + 1  
col_bm     <- (lipa_df[tip_order, "q.value.bm"] < 0.05) + 1

col_matrix <- cbind(col_growth, col_random, col_bm)


barplot.phylo4d(p4d_growth_rate, bar.col = col_matrix,  
                tree.type = "phylo", tree.ladderize = TRUE, 
                center = FALSE, scale = FALSE)
barplot.phylo4d(growth_rate.lipa.p4d, bar.col = col_matrix,  
                tree.type = "phylo", tree.ladderize = TRUE, 
                center = FALSE, scale = FALSE)
```
Pagel's Lambda
```{r, message=FALSE, warning=FALSE, echo=TRUE}
phyloSignal(p4d = p4d_growth_rate_ultra, methods = c("Lambda"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
growth_rate.crlg <- phyloCorrelogram(p4d_growth_rate_ultra, trait = "growth_rate")
random.crlg <- phyloCorrelogram(p4d_growth_rate_ultra, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_growth_rate_ultra, trait = "bm")

#plot(bm.crlg, main = "Brownian Motion Simulation")
#plot(random.crlg, main = "Random Trait (No Signal)")
#plot(growth_rate.crlg, main = "Phenotypic Trait: Growth Rate")
```

Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_growth_rate_ultra, "phylo")
trait_data <- p4d_growth_rate_ultra@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(42)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)



# Rank Order Plot 
rank_plots <- list()
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  sorted_null <- sort(res$null_dist)  # smallest λ first → largest λ last
  null_df <- data.frame(
    lambda = sorted_null,                    # Sorted λ values
    rank = 1:length(sorted_null),           # 1=lowest λ, N=highest λ
    above_obs = sorted_null >= res$obs_lambda  # Red if ≥ observed
  )
  
  # Store plot
  p <- ggplot(null_df, aes(x = rank, y = lambda, color = above_obs)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = c("TRUE" = "red"), 
                   labels = c("≥ Observed λ"), 
                   name = "Null values:") +
    geom_hline(yintercept = res$obs_lambda, linetype = "dashed", 
               color = "blue", size = 1.0) +
    labs(x = "Rank (lowest to highest λ)", y = "Pagel's λ statistic",
         title = trait_name,
         subtitle = sprintf("Observed λ = %.4f (p = %.4f)", 
                           res$obs_lambda, res$p_value)) +
    scale_x_continuous(expand = c(0.01, 0.01)) +
  theme_minimal() +
theme(
  legend.position = "top",
  legend.key.size = unit(0.4, "cm"),        
  legend.text = element_text(size = 7),   
  legend.title = element_text(size = 7),   
  plot.title = element_text(size = 9, hjust = 0.5, face = "bold"),
  plot.subtitle = element_text(size = 8, hjust = 0.5),
  axis.title = element_text(size = 9)
)
  
  rank_plots[[trait_name]] <- p
}

grid.arrange(grobs = rank_plots, ncol = 3, top = "Pagel's λ Permutation Test: Rank Order Plots")



```

## Carrying Capacity 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
# Permute the actual growth_rate values (scramble across tree tips)
random <- sample(cc_vector, size = length(tip_labels), replace = FALSE)
# Scale to match observed growth_rate moments
observed <- cc_vector[match(tip_labels, names(cc_vector))]
target_mean <- mean(observed)
target_sd <- sd(observed)


# Brownian Motion for Moran's I Test (Non-Ultrametric)
bm_raw <- rTraitCont(tr_subset, model = "BM")
bm <- (bm_raw - mean(bm_raw)) / sd(bm_raw) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm), sd(bm))

dat_carrying_capacity <- data.frame(
  carrying_capacity = cc_vector[match(tip_labels, names(cc_vector))],   
  random      = random,
  bm          = bm
  
)
rownames(dat_carrying_capacity) <- tip_labels
p4d_carrying_capacity <- phylo4d(tr_subset, dat_carrying_capacity)


# Brownian Motion Pagel's Lambda (Ultrametric Tree)
bm_raw_ultra <- rTraitCont(tr_subset_ultra, model = "BM")
bm_ultra <- (bm_raw_ultra - mean(bm_raw_ultra)) / sd(bm_raw_ultra) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm_ultra), sd(bm_ultra))

dat_carrying_capacity_ultra <- data.frame(
  carrying_capacity = cc_vector[match(tip_labels, names(cc_vector))],   
  random      = random,
  bm          = bm_ultra
  
)
rownames(dat_carrying_capacity_ultra) <- tip_labels

p4d_carrying_capacity_ultra <- phylo4d(tr_subset_ultra, dat_carrying_capacity_ultra)
# Trait Distribution
barplot.phylo4d(p4d_carrying_capacity_ultra, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_carrying_capacity_ultra, tree.type = "phylo", tree.ladderize = TRUE)
```

Moran's I
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_carrying_capacity, methods = c("I"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
carrying_capacity.crlg <- phyloCorrelogram(p4d_carrying_capacity, trait = "carrying_capacity")
random.crlg <- phyloCorrelogram(p4d_carrying_capacity, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_carrying_capacity, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(carrying_capacity.crlg, main = "Phenotypic Trait: Growth Rate")


#Local Indicator of Phylogenetic Association (LIPA)
carrying_capacity.lipa <- lipaMoran(p4d_carrying_capacity)
carrying_capacity.lipa.p4d <- lipaMoran(p4d_carrying_capacity, as.p4d = TRUE)

barplot.phylo4d(p4d_carrying_capacity, bar.col = (carrying_capacity.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)
barplot.phylo4d(carrying_capacity.lipa.p4d, bar.col = (carrying_capacity.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)

lipa_df <- as.data.frame(carrying_capacity.lipa)
print(lipa_df)
# Filter significant p-values (≤ 0.05)
sig_lipa <- lipa_df[lipa_df$p.value.carrying_capacity <= 0.05, ]
# Print table
print(sig_lipa)

# Nicer formatted table (requires knitr)
kable(sig_lipa, caption = "Significant LIPA Results (p ≤ 0.05)")



```
Pagel's Lambda
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_carrying_capacity_ultra, methods = c("Lambda"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
carrying_capacity.crlg <- phyloCorrelogram(p4d_carrying_capacity_ultra, trait = "carrying_capacity")
random.crlg <- phyloCorrelogram(p4d_carrying_capacity_ultra, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_carrying_capacity_ultra, trait = "bm")

#plot(bm.crlg, main = "Brownian Motion Simulation")
#plot(random.crlg, main = "Random Trait (No Signal)")
#plot(carrying_capacity.crlg, main = "Phenotypic Trait: Carrying Capacity")
```
Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_carrying_capacity_ultra, "phylo")
trait_data <- p4d_carrying_capacity_ultra@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(42)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)


# Rank Order Plot 
rank_plots <- list()
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  sorted_null <- sort(res$null_dist)  # smallest λ first → largest λ last
  null_df <- data.frame(
    lambda = sorted_null,                    # Sorted λ values
    rank = 1:length(sorted_null),           # 1=lowest λ, N=highest λ
    above_obs = sorted_null >= res$obs_lambda  # Red if ≥ observed
  )
  
  # Store plot
  p <- ggplot(null_df, aes(x = rank, y = lambda, color = above_obs)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = c("TRUE" = "red"), 
                   labels = c("≥ Observed λ"), 
                   name = "Null values:") +
    geom_hline(yintercept = res$obs_lambda, linetype = "dashed", 
               color = "blue", size = 1.0) +
    labs(x = "Rank (lowest to highest λ)", y = "Pagel's λ statistic",
         title = trait_name,
         subtitle = sprintf("Observed λ = %.4f (p = %.4f)", 
                           res$obs_lambda, res$p_value)) +
    scale_x_continuous(expand = c(0.01, 0.01)) +
  theme_minimal() +
theme(
  legend.position = "top",
  legend.key.size = unit(0.4, "cm"),        
  legend.text = element_text(size = 7),   
  legend.title = element_text(size = 7),   
  plot.title = element_text(size = 9, hjust = 0.5, face = "bold"),
  plot.subtitle = element_text(size = 8, hjust = 0.5),
  axis.title = element_text(size = 9)
)
  
  rank_plots[[trait_name]] <- p
}

grid.arrange(grobs = rank_plots, ncol = 3, top = "Pagel's λ Permutation Test: Rank Order Plots")


```


## % Biofilm Formation
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
# Permute the actual growth_rate values (scramble across tree tips)
random <- sample(pb_vector, size = length(tip_labels_02), replace = FALSE)
# Scale to match observed growth_rate moments
observed <- pb_vector[match(tip_labels_02, names(pb_vector))]
target_mean <- mean(observed)
target_sd <- sd(observed)


# Brownian Motion for Moran's I Test (Non-Ultrametric)
bm_raw <- rTraitCont(tr_subset_02, model = "BM")
bm <- (bm_raw - mean(bm_raw)) / sd(bm_raw) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm), sd(bm))

dat_percent_biofilm <- data.frame(
  percent_biofilm = pb_vector[match(tip_labels_02, names(pb_vector))],   
  random      = random,
  bm          = bm
)
rownames(dat_percent_biofilm) <- tip_labels_02
p4d_percent_biofilm <- phylo4d(tr_subset_02, dat_percent_biofilm)


# Brownian Motion Pagel's Lambda (Ultrametric Tree)
bm_raw_ultra <- rTraitCont(tr_subset_ultra_02, model = "BM")
bm_ultra <- (bm_raw_ultra - mean(bm_raw_ultra)) / sd(bm_raw_ultra) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm_ultra), sd(bm_ultra))

dat_percent_biofilm_ultra <- data.frame(
  percent_biofilm = pb_vector[match(tip_labels_02, names(pb_vector))],   
  random      = random,
  bm          = bm_ultra
)
rownames(dat_percent_biofilm_ultra) <- tip_labels_02

p4d_percent_biofilm_ultra <- phylo4d(tr_subset_ultra_02, dat_percent_biofilm_ultra)
# Trait Distribution
barplot.phylo4d(p4d_percent_biofilm_ultra, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_percent_biofilm_ultra, tree.type = "phylo", tree.ladderize = TRUE)
```

Moran's I
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_percent_biofilm, methods = c("I"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
percent_biofilm.crlg <- phyloCorrelogram(p4d_percent_biofilm, trait = "percent_biofilm")
random.crlg <- phyloCorrelogram(p4d_percent_biofilm, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_percent_biofilm, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(percent_biofilm.crlg, main = "Phenotypic Trait: Growth Rate")


# Local Indicator of Phylogenetic Association (LIPA)
percent_biofilm.lipa <- lipaMoran(p4d_percent_biofilm)
percent_biofilm.lipa.p4d <- lipaMoran(p4d_percent_biofilm, as.p4d = TRUE)

barplot.phylo4d(p4d_percent_biofilm, bar.col = (percent_biofilm.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)
barplot.phylo4d(percent_biofilm.lipa.p4d, bar.col = (percent_biofilm.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)

lipa_df <- as.data.frame(percent_biofilm.lipa)
print(lipa_df)

# Filter significant p-values (≤ 0.05)
sig_lipa <- lipa_df[lipa_df$p.value.percent_biofilm <= 0.05, ]
# Print table
print(sig_lipa)


lipa_df$q.value.percent_biofilm <- p.adjust(lipa_df$p.value.percent_biofilm, "BH")
lipa_df$q.value.random <- p.adjust(lipa_df$p.value.random, "BH")
lipa_df$q.value.bm    <- p.adjust(lipa_df$p.value.bm, "BH")


sig_lipa <- lipa_df[
  lipa_df$q.value.percent_biofilm <= 0.05 | 
  lipa_df$q.value.random <= 0.05 | 
  lipa_df$q.value.bm <= 0.05, 
]
# Print table
print(sig_lipa)
kable(sig_lipa, caption = "FDR adjusted p-value: Significant LIPA Results (q ≤ 0.05)")


tip_order <- match(percent_biofilm.lipa.p4d@label, rownames(lipa_df))
col_growth <- (lipa_df[tip_order, "q.value.percent_biofilm"] < 0.05) + 1
col_random <- (lipa_df[tip_order, "q.value.random"] < 0.05) + 1  
col_bm     <- (lipa_df[tip_order, "q.value.bm"] < 0.05) + 1

col_matrix <- cbind(col_growth, col_random, col_bm)


barplot.phylo4d(p4d_percent_biofilm, bar.col = col_matrix,  
                tree.type = "phylo", tree.ladderize = TRUE, 
                center = FALSE, scale = FALSE) 
barplot.phylo4d(percent_biofilm.lipa.p4d, bar.col = col_matrix,  
                tree.type = "phylo", tree.ladderize = TRUE, 
                center = FALSE, scale = FALSE)


```

Pagel's Lambda
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_percent_biofilm_ultra, methods = c("Lambda"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
percent_biofilm.crlg <- phyloCorrelogram(p4d_percent_biofilm_ultra, trait = "percent_biofilm")
random.crlg <- phyloCorrelogram(p4d_percent_biofilm_ultra, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_percent_biofilm_ultra, trait = "bm")

#plot(bm.crlg, main = "Brownian Motion Simulation")
#plot(random.crlg, main = "Random Trait (No Signal)")
#plot(percent_biofilm.crlg, main = "Phenotypic Trait: Percent Biofilm")
```

Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_percent_biofilm_ultra, "phylo")
trait_data <- p4d_percent_biofilm_ultra@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(42)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)



# Rank Order Plot 
rank_plots <- list()
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  sorted_null <- sort(res$null_dist)  # smallest λ first → largest λ last
  null_df <- data.frame(
    lambda = sorted_null,                    # Sorted λ values
    rank = 1:length(sorted_null),           # 1=lowest λ, N=highest λ
    above_obs = sorted_null >= res$obs_lambda  # Red if ≥ observed
  )
  
  # Store plot
  p <- ggplot(null_df, aes(x = rank, y = lambda, color = above_obs)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = c("TRUE" = "red"), 
                   labels = c("≥ Observed λ"), 
                   name = "Null values:") +
    geom_hline(yintercept = res$obs_lambda, linetype = "dashed", 
               color = "blue", size = 1.0) +
    labs(x = "Rank (lowest to highest λ)", y = "Pagel's λ statistic",
         title = trait_name,
         subtitle = sprintf("Observed λ = %.4f (p = %.4f)", 
                           res$obs_lambda, res$p_value)) +
    scale_x_continuous(expand = c(0.01, 0.01)) +
  theme_minimal() +
theme(
  legend.position = "top",
  legend.key.size = unit(0.4, "cm"),        
  legend.text = element_text(size = 7),   
  legend.title = element_text(size = 7),   
  plot.title = element_text(size = 9, hjust = 0.5, face = "bold"),
  plot.subtitle = element_text(size = 8, hjust = 0.5),
  axis.title = element_text(size = 9)
)
  
  rank_plots[[trait_name]] <- p
}

grid.arrange(grobs = rank_plots, ncol = 3, top = "Pagel's λ Permutation Test: Rank Order Plots")
```

## Safranin  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
# Permute the actual growth_rate values (scramble across tree tips)
random <- sample(sf_vector, size = length(tip_labels_02), replace = FALSE)
# Scale to match observed growth_rate moments
observed <- sf_vector[match(tip_labels_02, names(sf_vector))]
target_mean <- mean(observed)
target_sd <- sd(observed)


# Brownian Motion for Moran's I Test (Non-Ultrametric)
bm_raw <- rTraitCont(tr_subset_02, model = "BM")
bm <- (bm_raw - mean(bm_raw)) / sd(bm_raw) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm), sd(bm))

dat_safranin <- data.frame(
  safranin = sf_vector[match(tip_labels_02, names(sf_vector))],   
  random      = random,
  bm          = bm
  
)
rownames(dat_safranin) <- tip_labels_02
p4d_safranin <- phylo4d(tr_subset_02, dat_safranin)


# Brownian Motion Pagel's Lambda (Ultrametric Tree)
bm_raw_ultra <- rTraitCont(tr_subset_ultra_02, model = "BM")
bm_ultra <- (bm_raw_ultra - mean(bm_raw_ultra)) / sd(bm_raw_ultra) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm_ultra), sd(bm_ultra))

dat_safranin_ultra <- data.frame(
  safranin = sf_vector[match(tip_labels_02, names(sf_vector))],   
  random      = random,
  bm          = bm_ultra
  
)
rownames(dat_safranin_ultra) <- tip_labels_02

p4d_safranin_ultra <- phylo4d(tr_subset_ultra_02, dat_safranin_ultra)
# Trait Distribution
barplot.phylo4d(p4d_safranin_ultra, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_safranin_ultra, tree.type = "phylo", tree.ladderize = TRUE)
``` 

Moran's I
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_safranin, methods = c("I"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
safranin.crlg <- phyloCorrelogram(p4d_safranin, trait = "safranin")
random.crlg <- phyloCorrelogram(p4d_safranin, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_safranin, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(safranin.crlg, main = "Phenotypic Trait: Safranin")


#Local Indicator of Phylogenetic Association (LIPA)
safranin.lipa <- lipaMoran(p4d_safranin)
safranin.lipa.p4d <- lipaMoran(p4d_safranin, as.p4d = TRUE)

barplot.phylo4d(p4d_safranin, bar.col = (safranin.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)
barplot.phylo4d(safranin.lipa.p4d, bar.col = (safranin.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)

lipa_df <- as.data.frame(safranin.lipa)
print(lipa_df)
# Filter significant p-values (≤ 0.05)
sig_lipa <- lipa_df[lipa_df$p.value.safranin <= 0.05, ]
# Print table
print(sig_lipa)

# Nicer formatted table (requires knitr)
kable(sig_lipa, caption = "Significant LIPA Results (p ≤ 0.05)")
```

Pagel's Lambda
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_safranin_ultra, methods = c("Lambda"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
safranin.crlg <- phyloCorrelogram(p4d_safranin_ultra, trait = "safranin")
random.crlg <- phyloCorrelogram(p4d_safranin_ultra, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_safranin_ultra, trait = "bm")

#plot(bm.crlg, main = "Brownian Motion Simulation")
#plot(random.crlg, main = "Random Trait (No Signal)")
#plot(safranin.crlg, main = "Phenotypic Trait: Safranin")
```


Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_safranin_ultra, "phylo")
trait_data <- p4d_safranin_ultra@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(42)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)


# Rank Order Plot 
rank_plots <- list()
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  sorted_null <- sort(res$null_dist)  # smallest λ first → largest λ last
  null_df <- data.frame(
    lambda = sorted_null,                    # Sorted λ values
    rank = 1:length(sorted_null),           # 1=lowest λ, N=highest λ
    above_obs = sorted_null >= res$obs_lambda  # Red if ≥ observed
  )
  
  # Store plot
  p <- ggplot(null_df, aes(x = rank, y = lambda, color = above_obs)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = c("TRUE" = "red"), 
                   labels = c("≥ Observed λ"), 
                   name = "Null values:") +
    geom_hline(yintercept = res$obs_lambda, linetype = "dashed", 
               color = "blue", size = 1.0) +
    labs(x = "Rank (lowest to highest λ)", y = "Pagel's λ statistic",
         title = trait_name,
         subtitle = sprintf("Observed λ = %.4f (p = %.4f)", 
                           res$obs_lambda, res$p_value)) +
    scale_x_continuous(expand = c(0.01, 0.01)) +
  theme_minimal() +
theme(
  legend.position = "top",
  legend.key.size = unit(0.4, "cm"),        
  legend.text = element_text(size = 7),   
  legend.title = element_text(size = 7),   
  plot.title = element_text(size = 9, hjust = 0.5, face = "bold"),
  plot.subtitle = element_text(size = 8, hjust = 0.5),
  axis.title = element_text(size = 9)
)
  
  rank_plots[[trait_name]] <- p
}

grid.arrange(grobs = rank_plots, ncol = 3, top = "Pagel's λ Permutation Test: Rank Order Plots")


```

## MIC 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
tip_labels_03 <- names(mic_vector)
tr_subset_03 <- keep.tip(tr_subset, tip_labels_03)
tr_subset_ultra_03 <- keep.tip(tr_subset_ultra, tip_labels_03)

# Permute the actual growth_rate values (scramble across tree tips)
random <- sample(mic_vector, size = length(tip_labels_03), replace = FALSE)
# Scale to match observed growth_rate moments
observed <- mic_vector[match(tip_labels_03, names(mic_vector))]
target_mean <- mean(observed)
target_sd <- sd(observed)

# Brownian Motion for Moran's I Test (Non-Ultrametric)
bm_raw <- rTraitCont(tr_subset_03, model = "BM")
bm <- (bm_raw - mean(bm_raw)) / sd(bm_raw) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm), sd(bm))

dat_mic <- data.frame(
  mic = mic_vector[match(tip_labels_03, names(mic_vector))],   
  random      = random,
  bm          = bm
)
rownames(dat_mic) <- tip_labels_03
p4d_mic <- phylo4d(tr_subset_03, dat_mic)

# Brownian Motion Pagel's Lambda (Ultrametric Tree)
bm_raw_ultra <- rTraitCont(tr_subset_ultra_03, model = "BM")
bm_ultra <- (bm_raw_ultra - mean(bm_raw_ultra)) / sd(bm_raw_ultra) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm_ultra), sd(bm_ultra))

dat_mic_ultra <- data.frame(
  mic = mic_vector[match(tip_labels_03, names(mic_vector))],  
  random      = random,
  bm          = bm_ultra
)
rownames(dat_mic_ultra) <- tip_labels_03

p4d_mic_ultra <- phylo4d(tr_subset_ultra_03, dat_mic_ultra)
# Trait Distribution
barplot.phylo4d(p4d_mic_ultra, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_mic_ultra, tree.type = "phylo", tree.ladderize = TRUE)
```
Moran's I
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_mic, methods = c("I"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
mic.crlg <- phyloCorrelogram(p4d_mic, trait = "mic")
random.crlg <- phyloCorrelogram(p4d_mic, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_mic, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(mic.crlg, main = "Phenotypic Trait: MIC")

#Local Indicator of Phylogenetic Association (LIPA)
mic.lipa <- lipaMoran(p4d_mic)
mic.lipa.p4d <- lipaMoran(p4d_mic, as.p4d = TRUE)

barplot.phylo4d(p4d_mic, bar.col = (mic.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)
barplot.phylo4d(mic.lipa.p4d, bar.col = (mic.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)

lipa_df <- as.data.frame(mic.lipa)
print(lipa_df)
# Filter significant p-values (≤ 0.05)
sig_lipa <- lipa_df[lipa_df$p.value.mic <= 0.05, ]
# Print table
print(sig_lipa)

# Nicer formatted table (requires knitr)
kable(sig_lipa, caption = "Significant LIPA Results (p ≤ 0.05)")
```

Pagel's Lambda
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
phyloSignal(p4d = p4d_mic_ultra, methods = c("Lambda"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
mic.crlg <- phyloCorrelogram(p4d_mic_ultra, trait = "mic")
random.crlg <- phyloCorrelogram(p4d_mic_ultra, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_mic_ultra, trait = "bm")

#plot(bm.crlg, main = "Brownian Motion Simulation")
#plot(random.crlg, main = "Random Trait (No Signal)")
#plot(mic.crlg, main = "Phenotypic Trait: MIC")
```

Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_mic_ultra, "phylo")
trait_data <- p4d_mic_ultra@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(42)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)


# Rank Order Plot 
rank_plots <- list()
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  sorted_null <- sort(res$null_dist)  # smallest λ first → largest λ last
  null_df <- data.frame(
    lambda = sorted_null,                    # Sorted λ values
    rank = 1:length(sorted_null),           # 1=lowest λ, N=highest λ
    above_obs = sorted_null >= res$obs_lambda  # Red if ≥ observed
  )
  
  # Store plot
  p <- ggplot(null_df, aes(x = rank, y = lambda, color = above_obs)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = c("TRUE" = "red"), 
                   labels = c("≥ Observed λ"), 
                   name = "Null values:") +
    geom_hline(yintercept = res$obs_lambda, linetype = "dashed", 
               color = "blue", size = 1.0) +
    labs(x = "Rank (lowest to highest λ)", y = "Pagel's λ statistic",
         title = trait_name,
         subtitle = sprintf("Observed λ = %.4f (p = %.4f)", 
                           res$obs_lambda, res$p_value)) +
    scale_x_continuous(expand = c(0.01, 0.01)) +
  theme_minimal() +
theme(
  legend.position = "top",
  legend.key.size = unit(0.4, "cm"),        
  legend.text = element_text(size = 7),   
  legend.title = element_text(size = 7),   
  plot.title = element_text(size = 9, hjust = 0.5, face = "bold"),
  plot.subtitle = element_text(size = 8, hjust = 0.5),
  axis.title = element_text(size = 9)
)
  
  rank_plots[[trait_name]] <- p
}

grid.arrange(grobs = rank_plots, ncol = 3, top = "Pagel's λ Permutation Test: Rank Order Plots")


```
## Minimum pH
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(42)
# Permute the actual growth_rate values (scramble across tree tips)
random <- sample(ph_vector, size = length(tip_labels), replace = FALSE)

# Scale to match observed growth_rate moments
observed <- ph_vector[match(tip_labels, names(ph_vector))]
target_mean <- mean(observed)
target_sd <- sd(observed)



# Brownian Motion for Moran's I Test (Non-Ultrametric)
bm_raw <- rTraitCont(tr_subset, model = "BM")
bm <- (bm_raw - mean(bm_raw)) / sd(bm_raw) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm), sd(bm))

dat_pH_growth <- data.frame(
  pH_growth = ph_vector[match(tip_labels, names(ph_vector))],   
  random      = random,
  bm          = bm
  
)
rownames(dat_pH_growth) <- tip_labels
p4d_pH_growth <- phylo4d(tr_subset, dat_pH_growth)


# Brownian Motion Pagel's Lambda (Ultrametric Tree)
bm_raw_ultra <- rTraitCont(tr_subset_ultra, model = "BM")
bm_ultra <- (bm_raw_ultra - mean(bm_raw_ultra)) / sd(bm_raw_ultra) * target_sd + target_mean
# Verify same mean and std. as dataset 
sprintf("Observed: mean=%.3f, sd=%.3f", target_mean, target_sd)
sprintf("BM sim:   mean=%.3f, sd=%.3f", mean(bm_ultra), sd(bm_ultra))

dat_pH_growth_ultra <- data.frame(
  pH_growth = ph_vector[match(tip_labels, names(ph_vector))],   
  random      = random,
  bm          = bm_ultra
  
)
rownames(dat_pH_growth_ultra) <- tip_labels

p4d_pH_growth_ultra <- phylo4d(tr_subset_ultra, dat_pH_growth_ultra)
# Trait Distribution
barplot.phylo4d(p4d_pH_growth_ultra, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_pH_growth_ultra, tree.type = "phylo", tree.ladderize = TRUE)
```

Moran's I
```{r, message=FALSE, warning=FALSE, echo=TRUE}
phyloSignal(p4d = p4d_pH_growth, methods = c("I"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
pH_growth.crlg <- phyloCorrelogram(p4d_pH_growth, trait = "pH_growth")
random.crlg <- phyloCorrelogram(p4d_pH_growth, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_pH_growth, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(pH_growth.crlg, main = "Phenotypic Trait: Growth Rate")


#Local Indicator of Phylogenetic Association (LIPA)
pH_growth.lipa <- lipaMoran(p4d_pH_growth)
pH_growth.lipa.p4d <- lipaMoran(p4d_pH_growth, as.p4d = TRUE)

barplot.phylo4d(p4d_pH_growth, bar.col = (pH_growth.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)
barplot.phylo4d(pH_growth.lipa.p4d, bar.col = (pH_growth.lipa$p.value < 0.05) + 1, 
                center = FALSE, scale = FALSE, tree.type = "phylo", tree.ladderize = TRUE)

lipa_df <- as.data.frame(pH_growth.lipa)
print(lipa_df)
# Filter significant p-values (≤ 0.05)
sig_lipa <- lipa_df[lipa_df$p.value.pH_growth <= 0.05, ]
# Print table
print(sig_lipa)

# Nicer formatted table (requires knitr)
kable(sig_lipa, caption = "Significant LIPA Results (p ≤ 0.05)")
```
Pagel's Lambda
```{r, message=FALSE, warning=FALSE, echo=TRUE}
phyloSignal(p4d = p4d_pH_growth_ultra, methods = c("Lambda"))
# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
pH_growth.crlg <- phyloCorrelogram(p4d_pH_growth_ultra, trait = "pH_growth")
random.crlg <- phyloCorrelogram(p4d_pH_growth_ultra, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_pH_growth_ultra, trait = "bm")

#plot(bm.crlg, main = "Brownian Motion Simulation")
#plot(random.crlg, main = "Random Trait (No Signal)")
#plot(growth_rate.crlg, main = "Phenotypic Trait: Growth Rate")
```

Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_pH_growth_ultra, "phylo")
trait_data <- p4d_pH_growth_ultra@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(42)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)



# Rank Order Plot 
rank_plots <- list()
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  sorted_null <- sort(res$null_dist)  # smallest λ first → largest λ last
  null_df <- data.frame(
    lambda = sorted_null,                    # Sorted λ values
    rank = 1:length(sorted_null),           # 1=lowest λ, N=highest λ
    above_obs = sorted_null >= res$obs_lambda  # Red if ≥ observed
  )
  
  # Store plot
  p <- ggplot(null_df, aes(x = rank, y = lambda, color = above_obs)) +
    geom_point(size = 1.5, alpha = 0.8) +
    scale_color_manual(values = c("TRUE" = "red"), 
                   labels = c("≥ Observed λ"), 
                   name = "Null values:") +
    geom_hline(yintercept = res$obs_lambda, linetype = "dashed", 
               color = "blue", size = 1.0) +
    labs(x = "Rank (lowest to highest λ)", y = "Pagel's λ statistic",
         title = trait_name,
         subtitle = sprintf("Observed λ = %.4f (p = %.4f)", 
                           res$obs_lambda, res$p_value)) +
    scale_x_continuous(expand = c(0.01, 0.01)) +
  theme_minimal() +
theme(
  legend.position = "top",
  legend.key.size = unit(0.4, "cm"),        
  legend.text = element_text(size = 7),   
  legend.title = element_text(size = 7),   
  plot.title = element_text(size = 9, hjust = 0.5, face = "bold"),
  plot.subtitle = element_text(size = 8, hjust = 0.5),
  axis.title = element_text(size = 9)
)
  
  rank_plots[[trait_name]] <- p
}

grid.arrange(grobs = rank_plots, ncol = 3, top = "Pagel's λ Permutation Test: Rank Order Plots")



```

