---
title: "(Rooted Updated Tree) Testing Phylogenetic Signal in Phenotypic Data"
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggtree)
library(phylosignal)
library(broom)
library(rstatix)
library(adephylo)
library(ape)
library(phylobase)
library(formattable)
library(corrplot)
library(ggbeeswarm)
library(cowplot)
library(ggpubr)
library(grid)
library(gridGraphics)
library(ggplot2)
library(readr) 
library(dplyr)
library(tidyr)
library(patchwork)
library(phytools)
`%!in%` <- negate(`%in%`)
```

## Prepare data for analysis 
Import phenotype and phylogenetic tree data
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Import phylogenetic tree 
#tr <- read.tree("/Users/spencertoth/Desktop/research_ongoing/gardnerella_phylogenetic_signal/gardnerella_pathogenic_phenotypes_1_17_2026/RAxML_bestTree.gardTree")
tr <- read.tree("/Users/spencertoth/Desktop/research_ongoing/gardnerella_phylogenetic_signal/unrootedGardPhylogeny_withOutgroup.tre")
outgroup <- "QUERY___B_longum_51A" 
#tr_rooted <- root(tr, outgroup = "QUERY___B_longum_51A", resolve.root = TRUE)
#tr <- tr_rooted
#tr <- read.tree("/Users/spencertoth/Desktop/research_ongoing/gardnerella_phylogenetic_signal/rootedGardPhylogeny_noOutgroup.tre")
is.rooted(tr)
# Rename tree nodes to match strain names in strain-phenotype data 
numeric_tips <- tr$tip.label[tr$tip.label %in% c("101", "5-1")]
tr$tip.label[match(numeric_tips, tr$tip.label)] <- paste0("Gv", numeric_tips)
tr$tip.label[tr$tip.label == "14018"] <- "S14018"
tip_labels <- tr$tip.label
is.rooted(tr)
```

Average the phenotype values across technical replicates per strain  
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Growth Rates
growth_rate_data_file <- read_csv("../formatted_phenotype_data/growth_rate.csv")
car_cap_data_file <- read_csv("../formatted_phenotype_data/carrying_capacity.csv")
percent_biofilm_od_data_file  <- read_csv("../formatted_phenotype_data/percent_biofilm_od.csv")
safranin_od_data_file <- read_csv("../formatted_phenotype_data/safranin_od.csv")

growth_rate_data <- growth_rate_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_mSlope = mean(mSlope, na.rm = TRUE))
gr_vector <- setNames(growth_rate_data$mean_mSlope, growth_rate_data$Strain)

car_cap_data <- car_cap_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_maxOD = mean(maxOD, na.rm = TRUE))
cc_vector <- setNames(car_cap_data$mean_maxOD, car_cap_data$Strain)

percent_biofilm_od_data <- percent_biofilm_od_data_file %>%
  mutate(Strain = as.character(Strain),
    #     arcsq_mPercentOD = asin(sqrt(mPercentOD/100))) %>%
          mPercentOD = mPercentOD/100) %>%
  group_by(Strain) %>% 
  summarise(mean_mPercentOD = mean(mPercentOD, na.rm = TRUE))   
pb_vector <- setNames(percent_biofilm_od_data$mean_mPercentOD, percent_biofilm_od_data$Strain)

safranin_od_data <- safranin_od_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_mSafOD = mean(mSafOD, na.rm = TRUE))
sf_vector <- setNames(safranin_od_data$mean_mSafOD, safranin_od_data$Strain)
```
Rename complete-numeric strains  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Strain 14018 not present in: pb_vector, sf_vector (did not pass QC for these experiments)
vecs <- list(gr = gr_vector, cc = cc_vector)  
vecs[] <- lapply(vecs, function(v) {
  idx <- names(v) == "14018"   
  if(any(idx)) names(v)[idx] <- "S14018"   
  v
})

gr_vector <<- vecs$gr
cc_vector <<- vecs$cc
```

## Update: "phylosignal" R-package used to conduct phylogenetic signal tests (Moran's I and Pagel’s lambda) 
Subset phylogenetic tree to the strains with phenotype information 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Phenotypes with all 22 strains: carrying capacity, growth rate and minimum pH
data_tips <- intersect(names(gr_vector), tr$tip.label)  
data_tips <- union(data_tips, outgroup) # keep outgroup  
tr_subset  <- keep.tip(tr, data_tips)
tr_rooted  <- root(tr_subset, outgroup = outgroup, resolve.root = TRUE)
tr_subset <- drop.tip(tr_rooted, outgroup)
#tr_subset <- drop.tip(tr_subset, outgroup)

tip_labels <- tr_subset$tip.label
n_strains <- length(tip_labels) # 22

p <- ggtree(tr_subset) +
  geom_rootedge(rootedge = 0.05, colour = "black", linewidth = 1) +
  geom_tiplab(size = 4) +
  xlim(NA, 0.75)
p

# Phenotypes with 21 strains: safranin and % biofilm 
data_tips_02 <- intersect(names(sf_vector), tr$tip.label)  
data_tips_02 <- union(data_tips_02, outgroup) # keep outgroup  
tr_subset_02  <- keep.tip(tr, data_tips_02)
tr_rooted_02  <- root(tr_subset_02, outgroup = outgroup, resolve.root = TRUE)
tr_subset_02 <- drop.tip(tr_rooted_02, outgroup)

tip_labels_02 <- tr_subset_02$tip.label
n_strains_02 <- length(tip_labels_02) # 21
```

```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
tr_subset_ultra <- force.ultrametric(tr_subset, method = "nnls") # nnls or extend
tr_subset_ultra$edge.length[tr_subset_ultra$edge.length <= 0] <- 1e-6


p <- ggtree(tr_subset_ultra) +
  geom_rootedge(rootedge = 0.05, colour = "black", linewidth = 1) +
  geom_tiplab(size = 4) +
  xlim(NA, 0.75)
p

tr_subset_ultra_02 <- force.ultrametric(tr_subset_02, method = "nnls") # nnls or extend
tr_subset_ultra_02$edge.length[tr_subset_ultra_02$edge.length <= 0] <- 1e-6



``` 

Moran's I and Pagel's Lambda: Phylogenetic Signal along BM Gradient 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Distribution of Test Statistic as percentage of BM increases
set.seed(0)
phylosim <- phyloSim(tree = tr_subset_ultra, methods = c("I", "Lambda"), model = "BM", pb = FALSE)
```

Phylogenetic Signal Statistics Along BM-gradient
```{r, message=FALSE, warning=FALSE, echo=TRUE}
plot(phylosim, stacked.methods = FALSE, quantiles = c(0, 1), ylab = "Phylogenetic Signal Value")
plot.phylosim(phylosim, what = "pval", stacked.methods = TRUE)
mean(phylosim$sim.stat[1, 1, ])  # I at 0% BM 
mean(phylosim$sim.stat[1, 2, ])   # Lambda at 0% BM 

set.seed(0)
# Simulated traits along Brownian Motion given phylogenetic tree 
bm_raw <- rTraitCont(tr_subset_ultra, model = "BM")
# Normalize to 0-1 range for comparison with growth rate data 
bm <- (bm_raw - min(bm_raw)) / (max(bm_raw) - min(bm_raw))


set.seed(0)
# Simulated traits along Brownian Motion given phylogenetic tree 
bm_raw_02 <- rTraitCont(tr_subset_ultra_02, model = "BM")
# Normalize to 0-1 range for comparison with growth rate data 
bm_02 <- (bm_raw_02 - min(bm_raw_02)) / (max(bm_raw_02) - min(bm_raw_02))
```
Exponential Growth Rate
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(0)
# Random generated trait dataset with SD of observed dataset
random_raw <- rnorm(22, sd = 0.146)
# Scaled to 0-1 range for comparison with growth rate data
random <- (random_raw - min(random_raw)) / (max(random_raw) - min(random_raw))

dat_growth_rate <- data.frame(
  growth_rate = gr_vector[match(tip_labels, names(gr_vector))],   
  random      = random,
  bm          = bm
  
)
rownames(dat_growth_rate) <- tip_labels
p4d_growth_rate <- phylo4d(tr_subset_ultra, dat_growth_rate)
# Trait Distribution
barplot.phylo4d(p4d_growth_rate, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_growth_rate, tree.type = "phylo", tree.ladderize = TRUE)

phyloSignal(p4d = p4d_growth_rate, methods = c("I", "Lambda"))

# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
growth_rate.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "growth_rate")
random.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation (Strong Signal)")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(growth_rate.crlg, main = "Phenotypic Trait: Growth Rate")
            
```

Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_growth_rate, "phylo")
trait_data <- p4d_growth_rate@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(0)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)


```

Carrying Capacity 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(0)
# Random generated trait dataset 
random_raw <- rnorm(n_strains, sd = 0.113)
# Scaled to 0-1 range for comparison with growth rate data
random <- (random_raw - min(random_raw)) / (max(random_raw) - min(random_raw))

dat_carrying_capacity <- data.frame(
  carrying_capacity = cc_vector[match(tip_labels, names(cc_vector))],   
  random      = random,
  bm          = bm
  
)
rownames(dat_carrying_capacity) <- tip_labels
p4d_carrying_capacity <- phylo4d(tr_subset_ultra, dat_carrying_capacity)
# Trait Distribution
barplot.phylo4d(p4d_carrying_capacity, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_carrying_capacity, tree.type = "phylo", tree.ladderize = TRUE)

phyloSignal(p4d = p4d_carrying_capacity, methods = c("I", "Lambda"))

# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
growth_rate.crlg <- phyloCorrelogram(p4d_carrying_capacity, trait = "carrying_capacity")
random.crlg <- phyloCorrelogram(p4d_carrying_capacity, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_carrying_capacity, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation (Strong Signal)")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(growth_rate.crlg, main = "Phenotypic Trait: Growth Rate")
            
```

Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_carrying_capacity, "phylo")
trait_data <- p4d_carrying_capacity@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(0)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)


```



% Biofilm Formation
```{r, message=FALSE, warning=FALSE, echo=TRUE}
set.seed(0)
# Random generated trait dataset 
random_raw <- rnorm(21, sd = 0.232)
# Scaled to 0-1 range for comparison with growth rate data
random <- (random_raw - min(random_raw)) / (max(random_raw) - min(random_raw))


dat_percent_biofilm <- data.frame(
  percent_biofilm = pb_vector[match(tip_labels_02, names(pb_vector))],   
  random      = random,
  bm          = bm
  
)
rownames(dat_percent_biofilm) <- tip_labels_02
p4d_percent_biofilm <- phylo4d(tr_subset_ultra_02, dat_percent_biofilm)
# Trait Distribution
barplot.phylo4d(p4d_percent_biofilm, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_percent_biofilm, tree.type = "phylo", tree.ladderize = TRUE)

phyloSignal(p4d = p4d_percent_biofilm, methods = c("I", "Lambda"))

# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
growth_rate.crlg <- phyloCorrelogram(p4d_percent_biofilm, trait = "carrying_capacity")
random.crlg <- phyloCorrelogram(p4d_percent_biofilm, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_percent_biofilm, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation (Strong Signal)")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(growth_rate.crlg, main = "Phenotypic Trait: Growth Rate")
            
```
