---
title: "(Rooted Updated Tree) Testing Phylogenetic Signal in Phenotypic Data"
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggtree)
library(phylosignal)
library(broom)
library(rstatix)
library(adephylo)
library(ape)
library(phylobase)
library(formattable)
library(corrplot)
library(ggbeeswarm)
library(cowplot)
library(ggpubr)
library(grid)
library(gridGraphics)
library(ggplot2)
library(geiger)
library(readr) 
library(dplyr)
library(tidyr)
library(patchwork)
library(phytools)
library(vegan)
`%!in%` <- negate(`%in%`)
```

## Prepare data for analysis 
Import phenotype and phylogenetic tree data
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Import phenotypic data

# Growth Rates
growth_rate_data_file <- read_csv("../formatted_phenotype_data/growth_rate.csv")
car_cap_data_file <- read_csv("../formatted_phenotype_data/carrying_capacity.csv")

# Import phylogenetic tree 
#tr <- read.tree("/Users/spencertoth/Desktop/research_ongoing/gardnerella_phylogenetic_signal/gardnerella_pathogenic_phenotypes_1_17_2026/RAxML_bestTree.gardTree")
#tr <- read.tree("/Users/spencertoth/Desktop/research_ongoing/gardnerella_phylogenetic_signal/unrootedGardPhylogeny_withOutgroup.tre")
#outgroup <- "QUERY___B_longum_51A" 
#tr_rooted <- root(tr, outgroup = "QUERY___B_longum_51A", resolve.root = TRUE)
#tr <- tr_rooted
tr <- read.tree("/Users/spencertoth/Desktop/research_ongoing/gardnerella_phylogenetic_signal/rootedGardPhylogeny_noOutgroup.tre")
is.rooted(tr)
# Rename tree nodes to match strain names in strain-phenotype data 
numeric_tips <- tr$tip.label[tr$tip.label %in% c("101", "5-1")]
tr$tip.label[match(numeric_tips, tr$tip.label)] <- paste0("Gv", numeric_tips)
tr$tip.label[tr$tip.label == "14018"] <- "S14018"
tip_labels <- tr$tip.label
is.rooted(tr)
```

Average the phenotype values across technical replicates per strain  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
growth_rate_data <- growth_rate_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_mSlope = mean(mSlope, na.rm = TRUE))
gr_vector <- setNames(growth_rate_data$mean_mSlope, growth_rate_data$Strain)
gr_range <- range(gr_vector, na.rm = TRUE)

car_cap_data <- car_cap_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_maxOD = mean(maxOD, na.rm = TRUE))
cc_vector <- setNames(car_cap_data$mean_maxOD, car_cap_data$Strain)
cc_range <- range(cc_vector, na.rm = TRUE)
```
Rename complete-numeric strains  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Strain 14018 not present in: pb_vector, sf_vector (did not pass QC for these experiments)
vecs <- list(gr = gr_vector, cc = cc_vector)   
vecs[] <- lapply(vecs, function(v) {
  idx <- names(v) == "14018"   
  if(any(idx)) names(v)[idx] <- "S14018"   
  v
})

# Update global vectors
gr_vector <<- vecs$gr
cc_vector <<- vecs$cc

```

## Update: "phylosignal" R-package used to conduct phylogenetic signal tests (Moran's I and Pagel’s lambda) 
Subset phylogenetic tree to the strains with phenotype information 
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Phenotypes with all 22 strains: carrying capacity, growth rate and minimum pH
data_tips <- intersect(names(gr_vector), tr$tip.label)  
#data_tips <- union(data_tips, outgroup) # keep outgroup  
tr_subset  <- keep.tip(tr, data_tips)
#tr_rooted  <- root(tr_subset, outgroup = outgroup, resolve.root = TRUE)
#tr_subset <- drop.tip(tr_rooted, outgroup)
#tr_subset <- drop.tip(tr_subset, outgroup)

tip_labels <- tr_subset$tip.label
n_strains <- length(tip_labels) # 22

p <- ggtree(tr_subset) +
  geom_rootedge(rootedge = 0.05, colour = "black", linewidth = 1) +
  geom_tiplab(size = 4) +
  xlim(NA, 0.75)
p
```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Function to get root-to-tip distances for all tips
get_root_to_tip <- function(tree) {
  # For a rooted tree, the root is usually node Ntip + 1
  root_node <- Ntip(tree) + 1
  
  # All pairwise node distances
  D <- dist.nodes(tree)
  
  # Tips are rows 1..Ntip
  tip_indices <- 1:Ntip(tree)
  root_to_tip <- D[tip_indices, root_node]
  
  # Return as a data frame
  data.frame(
    tip = tree$tip.label,
    root_to_tip_distance = as.numeric(root_to_tip),
    row.names = NULL
  )
}

root_to_tip_table <- get_root_to_tip(tr_subset)
print(root_to_tip_table)

p <- ggplot(root_to_tip_table, aes(x = root_to_tip_distance)) +
  geom_histogram(bins = 22, fill = "steelblue", color = "black", alpha = 0.7) +
  xlim(0, 0.85) +
  ylim(0, 7) +
  labs(
    title = "Root-to-Tip Path Distances",
    x = "Distance to Root",
    y = "Count",
    subtitle = paste("N =", nrow(root_to_tip_table), "tips")
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

print(p)

# Summary stats
summary(root_to_tip_table$root_to_tip_distance)

```

Moran's I and Pagel's Lambda: Phylogenetic Signal along BM Gradient 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Distribution of Test Statistic as percentage of BM increases
set.seed(42)

plot(tr_subset, type="phylogram", root.edge=0.5)
is.rooted(tr_subset)
phylosim <- phyloSim(tree = tr_subset, methods = c("I", "Lambda"), model = "BM", pb = FALSE)

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
plot(phylosim, stacked.methods = FALSE, quantiles = c(0, 1), ylab = "Phylogenetic Signal Value")
plot.phylosim(phylosim, what = "pval", stacked.methods = TRUE)
mean(phylosim$sim.stat[1, 1, ])  # I at 0% BM 
mean(phylosim$sim.stat[1, 2, ])   # Lambda at 0% BM 

is.rooted(tr_subset)
set.seed(42)
# Simulated traits along Brownian Motion given phylogenetic tree 
bm_raw <- rTraitCont(tr_subset, model = "BM")
# Normalize to 0-1 range for comparison with growth rate data 
bm <- (bm_raw - min(bm_raw)) / (max(bm_raw) - min(bm_raw))

# Random generated trait dataset 
random_raw <- rnorm(n_strains)
# Scaled to 0-1 range for comparison with growth rate data
random <- (random_raw - min(random_raw)) / (max(random_raw) - min(random_raw))

dat_growth_rate <- data.frame(
  growth_rate = gr_vector[match(tip_labels, names(gr_vector))],   
  random      = random,
  bm          = bm
  
)

dat_carrying_capacity <- data.frame(
  carrying_capacity = cc_vector[match(tip_labels, names(cc_vector))],   
  random      = random,
  bm          = bm
  
)

rownames(dat_growth_rate) <- tip_labels

p4d_growth_rate <- phylo4d(tr_subset, dat_growth_rate)
p4d_carrying_capacity <- phylo4d(tr_subset, dat_carrying_capacity)

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Trait Distribution
barplot.phylo4d(p4d_growth_rate, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_growth_rate, tree.type = "phylo", tree.ladderize = TRUE)

phyloSignal(p4d = p4d_growth_rate, methods = c("I", "Lambda"))

# Phylogenetic correlogram (uses Local Moran's I autocorrelation)
growth_rate <- phyloCorrelogram(p4d_growth_rate, trait = "growth_rate")
random.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation (Strong Signal)")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(growth_rate.crlg, main = "Phenotypic Trait: growth_rate")

# Local Indicator of Phylogenetic Association (LIPA)
growth_rate.lipa <- lipaMoran(p4d_growth_rate)
growth_rate.lipa.p4d <- lipaMoran(p4d_growth_rate, as.p4d = TRUE)

barplot.phylo4d(p4d_growth_rate, bar.col = (growth_rate.lipa$p.value < 0.05) + 1, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
barplot.phylo4d(growth_rate.lipa.p4d, bar.col = (growth_rate.lipa$p.value < 0.05) + 1, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
```
Identify model of trait evolution with best fit to the phenotypic data
```{r, message=FALSE, warning=FALSE, echo=TRUE}

phy <- as(p4d_growth_rate, "phylo")
tip_data <- tdata(p4d_growth_rate, "tip")   

# Convert to named vectors (fitContinuous needs this)
dat_growth <- setNames(tip_data$growth_rate, rownames(tip_data))
dat_random <- setNames(tip_data$random, rownames(tip_data))
dat_bm     <- setNames(tip_data$bm, rownames(tip_data))


phy <- as(p4d_carrying_capacity, "phylo")
tip_data <- tdata(p4d_carrying_capacity, "tip")   
# Convert to named vectors (fitContinuous needs this)
dat_cc <- setNames(tip_data$carrying_capacity, rownames(tip_data))




# Verify names match tree
stopifnot(identical(names(dat_growth), phy$tip.label))

run_model_comparison <- function(phy, dat_vec, trait_name) {
  models <- c("BM", "OU", "white")
  aic_vals <- numeric(length(models)); names(aic_vals) <- models
  
  for (m in models) {
    fit <- fitContinuous(phy, dat_vec, model = m, ncores = 2)
    aic_vals[m] <- fit$opt$aic
  }
  
  weights <- aicw(aic_vals)
  weight_col <- weights[, 3]   
  max_wt <- max(weight_col)
  best_model <- rownames(weights)[which.max(weight_col)]
  
  cat("\n---", trait_name, "---\n")
  print(weights)
  
  if (max_wt >= 0.5) {
  if (best_model == "white") {   
    best_model <- "white_noise"
  }
  cat("✓", best_model, "better (wt =", round(max_wt, 3), ")\n")
} else {
    cat("No ideal model\n")
  }
  return(weights)
}

results_growth <- run_model_comparison(phy, dat_growth, "growth_rate")
results_growth <- run_model_comparison(phy, dat_cc, "carrying_capacity")
results_random <- run_model_comparison(phy, dat_random, "random") 
results_bm     <- run_model_comparison(phy, dat_bm, "bm")


``` 

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Master function for complete phylogenetic signal analysis
analyze_phylo4d <- function(p4d_obj_name) {
  # Get object dynamically
  p4d_obj <- get(p4d_obj_name)
  trait_names <- colnames(tdata(p4d_obj, "tip"))
  
  cat("\n=== ANALYZING:", p4d_obj_name, "===\n")
  cat("Traits:", paste(trait_names, collapse = ", "), "\n\n")
  
  # 1. Barplots (raw + Z-scores)
  barplot.phylo4d(p4d_obj, tree.type = "phylo", tree.ladderize = TRUE, 
                  center = FALSE, scale = FALSE)
  title(main = paste(p4d_obj_name, "- Raw Values"))
  
  barplot.phylo4d(p4d_obj, tree.type = "phylo", tree.ladderize = TRUE)
  title(main = paste(p4d_obj_name, "- Z-scores"))
  
  # 2. Phylogenetic signal (I, Lambda)
  signal_results <- phyloSignal(p4d = p4d_obj, methods = c("I", "Lambda"))
  print(signal_results)
  
  # 3. Correlograms for each trait
  crlg_results <- list()
  for(trait in trait_names) {
    crlg <- phyloCorrelogram(p4d_obj, trait = trait)
    crlg_results[[trait]] <- crlg
    plot(crlg, main = paste(trait, "Correlogram"))
  }
  
  # 4. LIPA analysis
  lipa_results <- lipaMoran(p4d_obj)
  lipa_p4d <- lipaMoran(p4d_obj, as.p4d = TRUE)
  
  # 5. LIPA-colored barplots
  barplot.phylo4d(p4d_obj, 
                  bar.col = (lipa_results$p.value < 0.05) + 1, 
                  tree.type = "phylo", tree.ladderize = TRUE, 
                  center = FALSE, scale = FALSE)
  title(main = paste(p4d_obj_name, "- LIPA p<0.05"))
  
  barplot.phylo4d(lipa_p4d, 
                  bar.col = (lipa_results$p.value < 0.05) + 1, 
                  tree.type = "phylo", tree.ladderize = TRUE, 
                  center = FALSE, scale = FALSE)
  
  # Return results
  return(list(
    phylo4d = p4d_obj,
    signal = signal_results,
    correlograms = crlg_results,
    lipa = lipa_results
  ))
}



# Usage - works for ANY p4d_* object:
results_growth <- analyze_phylo4d("p4d_growth_rate")
results_cc     <- analyze_phylo4d("p4d_carrying_capacity")   

# Quick one-liner for single trait focus:
analyze_trait <- function(p4d_obj_name, trait_name) {
  p4d_obj <- get(p4d_obj_name)
  cat("Trait:", trait_name, "in", p4d_obj_name, "\n")
  
  # Correlogram + LIPA only
  crlg <- phyloCorrelogram(p4d_obj, trait = trait_name)
  lipa <- lipaMoran(p4d_obj)
  
  plot(crlg, main = paste(trait_name, "- Correlogram"))
  print(lipa)
}

```
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Shared phylogenetic distance matrix
phy <- as(p4d_growth_rate, "phylo")  # Same tree for both
phy_dist <- cophenetic(phy)

# Extract BOTH traits into single matrix (rows=tips, cols=traits)
growth_data <- tdata(p4d_growth_rate, "tip")
cc_data <- tdata(p4d_carrying_capacity, "tip")


# Combine traits
all_traits <- cbind(
  growth_rate = growth_data$growth_rate,
  carrying_capacity = cc_data$carrying_capacity  # Adjust column name if needed
)
rownames(all_traits) <- rownames(growth_data)

# Multivariate Mantel (traits combined)
multivariate_dist <- dist(all_traits)  # Euclidean distance across both traits
mantel_multi <- mantel(phy_dist, multivariate_dist, permutations = 999)

cat("\nMultivariate Mantel (growth + cc):\n")
cat("r =", round(mantel_multi$statistic, 3), "p =", round(mantel_multi$signif, 3), "\n")

# Visualization
#par(mfrow = c(2,2))
#plot(phy_dist[upper.tri(phy_dist)], growth_dist[upper.tri(growth_dist)], 
 #    main = "Growth Rate", xlab = "Phylogeny", ylab = "Trait Dist")
#abline(0,1, col = "red", lty = 2)

#plot(phy_dist[upper.tri(phy_dist)], cc_dist[upper.tri(cc_dist)], 
 #    main = "Carrying Capacity", xlab = "Phylogeny", ylab = "Trait Dist")
#abline(0,1, col = "red", lty = 2)
```




Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_growth_rate, "phylo")
trait_data <- p4d_growth_rate@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(42)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
   
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)


```
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))

for (trait_name in names(results)) {
  res <- results[[trait_name]]
  null_vals <- res$null_dist
  obs <- res$obs_lambda
  
  # empirical p-value as proportion of null >= observed
  p_emp <- mean(null_vals >= obs)
  
  # base histogram of null λ (counts)
  h_all <- hist(null_vals,
                breaks = 30,
                plot = FALSE)
  
  # histogram of tail nulls (for over-plotting)
  tail_vals <- null_vals[null_vals >= obs]
  h_tail <- hist(tail_vals,
                 breaks = h_all$breaks,
                 plot = FALSE)
  
  # plot all nulls
  plot(h_all,
       xlim = c(0, 1),
       xlab = "λ",
       ylab = "Count",
       main = paste0(trait_name, " (p = ", signif(p_emp, 3), ")"),
       col = "grey85",
       border = "grey40")
  
  # overlay tail counts as shaded bars
  rect(xleft = h_tail$mids - 0.5 * diff(h_all$breaks)[1],
       ybottom = 0,
       xright = h_tail$mids + 0.5 * diff(h_all$breaks)[1],
       ytop = h_tail$counts,
       col = rgb(1, 0, 0, 0.4),
       border = NA)
  
  # vertical line for observed λ
  abline(v = obs, col = "red", lwd = 3, lty = 2)
}

# Print Results
results_df <- data.frame(
  trait   = trait_names,
  lambda  = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)
``` 

(Ultrametric Test)  Forced-Ultrametric Tree --> Simulated Brownian-Motion (BM) model (Phylogeny with 22 strains)
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Distribution of Test Statistic as percentage of BM increases
set.seed(42)
tr_subset_ultra <- force.ultrametric(tr_subset, method = "nnls") # nnls or extend
plot(tr_subset_ultra)
tr_subset_ultra$edge.length[tr_subset_ultra$edge.length <= 0] <- 1e-6


p <- ggtree(tr_subset_ultra) +
  geom_rootedge(rootedge = 0.05, colour = "black", linewidth = 1) +
  geom_tiplab(size = 4) +
  xlim(NA, 0.75)
p

phylosim <- phyloSim(tree = tr_subset_ultra, methods = c("I", "Lambda"), model = "BM", pb = FALSE)



root_to_tip_table <- get_root_to_tip(tr_subset_ultra)
print(root_to_tip_table)

p <- ggplot(root_to_tip_table, aes(x = root_to_tip_distance)) +
  geom_histogram(bins = 22, fill = "steelblue", color = "black", alpha = 0.7) +
  #xlim(0, 0.85) +
 # ylim(0, 7) +
  labs(
    title = "Root-to-Tip Path Distances (tr_subset_ultra)",
    x = "Distance to Root",
    y = "Count",
    subtitle = paste("N =", nrow(root_to_tip_table), "tips")
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

p

# Summary stats
summary(root_to_tip_table$root_to_tip_distance)
```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
mean(phylosim$sim.stat[1, 1, ])  # I at 0% BM 
mean(phylosim$sim.stat[1, 2, ])   # Lambda at 0% BM 

plot(phylosim, stacked.methods = FALSE, quantiles = c(0, 1), ylab = "Phylogenetic Signal Value")
plot.phylosim(phylosim, what = "pval", stacked.methods = TRUE)

set.seed(42)
# Simulated traits along Brownian Motion given phylogenetic tree 
bm_raw <- rTraitCont(tr_subset_ultra, model = "BM")
# Normalize to 0-1 range for comparison with growth rate data 
bm <- (bm_raw - min(bm_raw)) / (max(bm_raw) - min(bm_raw))

dat_growth_rate <- data.frame(
  growth_rate = gr_vector[match(tip_labels, names(gr_vector))],   
  random      = random,
  bm          = bm
  
)
rownames(dat_growth_rate) <- tip_labels
p4d_growth_rate <- phylo4d(tr_subset_ultra, dat_growth_rate)
```

Identify model of trait evolution with best fit to the phenotypic data
```{r, message=FALSE, warning=FALSE, echo=TRUE}

phy <- as(p4d_growth_rate, "phylo")
tip_data <- tdata(p4d_growth_rate, "tip")   

# Convert to named vectors (fitContinuous needs this)
dat_growth <- setNames(tip_data$growth_rate, rownames(tip_data))
dat_random <- setNames(tip_data$random, rownames(tip_data))
dat_bm     <- setNames(tip_data$bm, rownames(tip_data))

# Verify names match tree
stopifnot(identical(names(dat_growth), phy$tip.label))

cat("Ultrametric tree: model analysis")
results_growth <- run_model_comparison(phy, dat_growth, "growth_rate")
results_random <- run_model_comparison(phy, dat_random, "random") 
results_bm     <- run_model_comparison(phy, dat_bm, "bm")

``` 

```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Trait Distribution
barplot.phylo4d(p4d_growth_rate, tree.type = "phylo", tree.ladderize = TRUE, center = FALSE, scale = FALSE)
# Z-scores 
barplot.phylo4d(p4d_growth_rate, tree.type = "phylo", tree.ladderize = TRUE)

phyloSignal(p4d = p4d_growth_rate, methods = c("I", "Lambda"))

growth_rate.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "growth_rate")
random.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "random")
bm.crlg <- phyloCorrelogram(p4d_growth_rate, trait = "bm")

plot(bm.crlg, main = "Brownian Motion Simulation (Strong Signal)")
plot(random.crlg, main = "Random Trait (No Signal)")
plot(growth_rate.crlg, main = "Phenotypic Trait: Growth Rate")
```

```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
tr <- as(p4d_growth_rate, "phylo")
trait_data <- tdata(p4d_growth_rate, type="tip")
vcv <- vcv.phylo(tr)   

lambdaTest_fixed <- function(x, vcv, lambda_fixed) {
  logL0 <- pagelLogLik(0, x, vcv)  # Null: λ=0
  logL_fixed <- pagelLogLik(lambda_fixed, x, vcv)  # λ=fixed
  LR <- 2 * (logL_fixed - logL0)
  pvalue <- pchisq(LR, df=1, lower.tail=FALSE)
  return(list(lambda=lambda_fixed, logL_null=logL0, logL_alt=logL_fixed, LR=LR, pvalue=pvalue))
}

gr_lrt1 <- lambdaTest_fixed(trait_data$growth_rate, vcv, 1.0175)
cat("p-value =", gr_lrt1$pvalue, "\n")
```
Perform permutation test on Pagel's Lambda statistic 
```{r, message=FALSE, warning=FALSE, echo=TRUE, results="hide"}
# Extract tree + observed and simulated traits  
tree <- as(p4d_growth_rate, "phylo")
trait_data <- p4d_growth_rate@data  # (growth_rate, random, bm)
trait_names <- colnames(trait_data)

# Number of permutations
n_perm <- 1000
set.seed(42)

# Store results for all traits
results <- list()

# Loop through each trait
for(j in 1:ncol(trait_data)) {
  trait <- trait_data[[j]]
  trait_name <- trait_names[j]
  
  cat("\n---", trait_name, "---\n")
  
 # Pagel's λ  
  obs_lambda <- phylosig(tree, trait, method = "lambda")$lambda
  
  # Generate null distribution using 1000 permutations of trait data
  lambda_null <- numeric(n_perm)
  for(i in 1:n_perm) {
    shuffled_trait <- sample(trait)
    lambda_null[i] <- phylosig(tree, shuffled_trait, method = "lambda")$lambda
  }
  
  # right-tailed (We reject H₀ if observed λ is unexpectedly large i.e. right tail)
  p_val <- mean(lambda_null >= obs_lambda)
  
  results[[trait_name]] <- list(
    obs_lambda = obs_lambda,
    p_value = p_val,
    null_dist = lambda_null
  )
  
  cat("Observed λ =", round(obs_lambda, 4), "\n")
  cat("p-value =", round(p_val, 4), "\n")
}

```

```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Plot null distribution 
par(mfrow=c(1,3), mar=c(4,4,2,1))
for(trait_name in names(results)) {
  res <- results[[trait_name]]
  dens <- density(res$null_dist)
  
  plot(dens, main=trait_name, xlab="λ", lwd=2, 
       xlim=c(0,1), ylim=c(0,max(dens$y)*1.2))
  abline(v=res$obs_lambda, col="red", lwd=3, lty=2)
  
  # Shade area beyond observed
  x_tail <- seq(res$obs_lambda, 1, len=100)
  y_tail <- approx(dens$x, dens$y, xout=x_tail)$y
  polygon(c(res$obs_lambda, x_tail, 1), c(0, y_tail, 0), 
          col=rgb(1,0,0,0.3), border=NA)
}

# Print Results 
results_df <- data.frame(
  trait = trait_names,
  lambda = sapply(results, function(x) round(x$obs_lambda, 4)),
  p_value = sapply(results, function(x) round(x$p_value, 4))
)
print(results_df)


```
