---
title: "Testing Phylogenetic Signal in Phentypic Data"
output: html_document
---

**Purpose:**
  To assess phylogenetic association of *Gardnerella* phenotypes related to pathogenicity.

  Packages
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ape)
library(rstatix)

`%!in%` <- negate(`%in%`)
```

  Phenotype data
```{r}
# Growth Rates
growth_rate_DF <- read_csv("../formatted_phenotype_data/growth_rate.csv")
carrying_capacity_DF <- read_csv("../formatted_phenotype_data/carrying_capacity.csv")

# Biofilm formation
percent_biofilm_od_DF <- read_csv("../formatted_phenotype_data/percent_biofilm_od.csv")
safranin_od_DF <- read_csv("../formatted_phenotype_data/safranin_od.csv")

# Growth over pH
pH_growth_DF <- read_csv("../formatted_phenotype_data/pH_growth.csv")

# Metronidazole pH
metro_mic_DF <- read_csv("../metronidazole_mic/data/strainMICs.csv")
```

  Short cuts
```{r}
strains <- c("14018", "315-A", "C0011E4", "JCP7275", "C0084H9", "JCP8108", "JCP8017B", "C0040C2" , "UM35",  "JCP8066", "C0093B3", "AMD", "C0096A1", "C0102A2",  "UM224", "Gv5-1", "C0179B4", "C0056B5", "Gv101", "C0100B2", "C0101A1", "CMW7778B")

strains0 <- c("14018", "315-A", "C0011E4", "JCP7275", "C0084H9", "JCP8108", "JCP8017B", "C0040C2" , "UM35",  "JCP8066", "C0093B3", "AMD", "C0096A1", "C0102A2",  "UM224", "5-1", "C0179B4", "C0056B5", "101", "C0100B2", "C0101A1", "CMW7778B")

species <- c("Gardnerella vaginalis", "Gardnerella sp. 2", "Gardnerella sp. 3", "Gardnerella piotii", "Gardnerella leopoldii", "Gardnerella swidsinskii", "Gardnerella sp. 7", "Gardnerella sp. 8", "Gardnerella sp. 10", "Gardnerella sp. 11", "Gardnerella sp. 12")
```


  Phylogeny
```{r}
tree_path <- "../strain_phylogeny/RAxML_bestTree.gardTree"
tr <- read.tree(tree_path)
```


# Moran's I
  Choose the weights as $_{wij} = 1/d_{ij}$ , where the d’s is the distances measured on the tree:
```{r}
w <- 1/cophenetic(tr)
diag(w) <- 0 # set diagonals to zero
```

```{r}
W <- w[rownames(w) %in% strains0 ,colnames(w) %in% strains0]
W
```

## Growth Rates
### Exponential Growth Rate
```{r}
growthRates <- growth_rate_DF %>%
  with_groups(Strain, summarize, x=mean(mSlope)) %>%
  mutate(Strain=as.character(Strain),
         Strain=case_when(Strain=="Gv101"~"101", # fix differences between these strain name styles
                          Strain=="Gv5-1"~"5-1",
                          Strain %!in% c("101", "5-1")~Strain),
         Strain=factor(Strain, levels=strains0)) %>%
  .$x
names(growthRates) <- strains0

# perfrm Moran's I
Moran.I(growthRates, W)
```

### Carrying Capacity
### Moran's I
```{r}
carCapacities <- carrying_capacity_DF %>%
  with_groups(Strain, summarize, x=mean(maxOD)) %>%
  mutate(Strain=as.character(Strain),
         Strain=case_when(Strain=="Gv101"~"101",
                          Strain=="Gv5-1"~"5-1",
                          Strain %!in% c("101", "5-1")~Strain),
         Strain=factor(Strain, levels=strains0)) %>%
  .$x
names(carCapacities) <- strains0

# perfrm Moran's I
Moran.I(carCapacities, W)
```



## Biofilm Formation
### Percent Biofilm Optical Density
```{r}
biofilmStrains <- strains0[2:22] # 14018 did not pass quality checks

odVals <- percent_biofilm_od_DF %>%
  with_groups(Strain, summarize, x=mean(mPercentOD)) %>%
  mutate(Strain=as.character(Strain),
         Strain=case_when(Strain=="Gv101"~"101",
                          Strain=="Gv5-1"~"5-1",
                          Strain %!in% c("101", "5-1")~Strain),
         Strain=factor(Strain, levels=biofilmStrains)) %>%
  .$x
names(odVals) <- biofilmStrains

# remove 14018 from W matrix
biofilmW <- w[rownames(w) %in% biofilmStrains ,colnames(w) %in% biofilmStrains]

#perform Moran's I
Moran.I(odVals, biofilmW) 
```

### Biofilm by Safranin Stain
```{r}
safVals <- safranin_od_DF %>%
 with_groups(Strain, summarize, x=mean(mSafOD)) %>%
 mutate(Strain=as.character(Strain),
        Strain=case_when(Strain=="Gv101"~"101",
                         Strain=="Gv5-1"~"5-1",
                         Strain %!in% c("101", "5-1")~Strain),
        Strain=factor(Strain, levels=biofilmStrains)) %>%
 .$x
names(safVals) <- biofilmStrains

#perform Moran's I
Moran.I(safVals, biofilmW) 
```

## pH growth
### Minimim pH with growth
```{r}
#Perform ttest to determine lowest pH with non-zero growth
pH_ttest <- pH_growth_DF %>%
  group_by(Species, Strain, pH) %>%
  t_test(mOD~1, mu=0.01, alternative = "greater") %>%
  add_significance("p")

#get minimum pH values with growth
minpHGrowth0 <- pH_ttest %>%
   group_by(Strain, Species) %>%
   filter(p.signif!="ns") %>%
   filter(pH==min(pH)) %>%
   mutate(Strain=as.character(Strain),
          Strain=case_when(Strain=="Gv101"~"101",
                           Strain=="Gv5-1"~"5-1",
                           Strain %!in% c("101", "5-1")~Strain),
          Strain=factor(Strain, levels=strains0),
          pH=as.numeric(pH))
 
minpHGrowth <- set_names(minpHGrowth0$pH, minpHGrowth0$Strain)
minpHGrowth

Moran.I(minpHGrowth, W)
```

## Metrnidazole_MIC
  not previously completed

# Session Info
```{r}
sessionInfo()
```