---
title: "Phenotype Distribution of Gardnerella spp. Strains"
author: "Spencer Toth"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document: default
---
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(phylosignal)
library(broom)
library(rstatix)
library(adephylo)
library(ape)
library(phylobase)
library(formattable)
library(corrplot)
library(ggbeeswarm)
library(cowplot)
library(ggpubr)
library(grid)
library(gridGraphics)
library(ggplot2)
library(readr) 
library(dplyr)
library(tidyr)
library(patchwork)
`%!in%` <- negate(`%in%`)
```

# Import and Prepare input files 
## Import files and edit strain names 
```{r, message=FALSE, warning=FALSE, echo=TRUE}

strainIDs <- read.csv("/Users/spencertoth/Desktop/Research/gardnerella_phylogenetic_signal/gardnerella_pathogenic_phenotypes/strain_phylogeny/strainIDs.csv", stringsAsFactors = FALSE)
strainIDs$strain <- as.character(strainIDs$strain)

# Growth Rates
carrying_capacity_data_file <- read_csv("../formatted_phenotype_data/carrying_capacity.csv")
growth_rate_data_file <- read_csv("../formatted_phenotype_data/growth_rate.csv")

# Biofilm formation
percent_biofilm_od_data_file  <- read_csv("../formatted_phenotype_data/percent_biofilm_od.csv")
safranin_od_data_file <- read_csv("../formatted_phenotype_data/safranin_od.csv")

# Minimum pH Growth
pH_growth_data_file <- read_csv("../formatted_phenotype_data/pH_growth.csv")

# Metronidazole MIC
metro_mic_data_file <- read.csv("../formatted_phenotype_data/metro_mic.csv")


#Update names with genomospecies 
update_strain_gs <- function(df) {
  df %>%
    mutate(Strain = as.character(Strain)) %>%
    left_join(select(strainIDs, strain, genomospecies), by = c("Strain" = "strain")) %>%
    mutate(Strain = paste0(Strain, "_", genomospecies)) %>%
    select(-any_of(c("genomospecies", "strain")))
}

# Carrying Capacity 
carrying_capacity_data_file <- update_strain_gs(carrying_capacity_data_file)
#n_strains <- length(unique(carrying_capacity_data_file$Strain))
#n_observations <-length(carrying_capacity_data_file$maxOD)

carrying_capacity_means <- carrying_capacity_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>%
  summarise(n_isolates = n(), mean_maxOD = mean(maxOD, na.rm = TRUE), .groups = "drop") %>%
  arrange(mean_maxOD) %>%
  mutate(Strain = factor(Strain, levels = Strain)) %>%
  mutate(n_strains = length(unique(Strain))) 


#Growth Rate
growth_rate_data_file <- update_strain_gs(growth_rate_data_file)

#n_strains <- length(unique(growth_rate_data_file$Strain))
#n_observations <-length(growth_rate_data_file$mSlope)

growth_rate_means <- growth_rate_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>%
  summarise(n_isolates = n(),mean_growth_rate = mean(mSlope, na.rm = TRUE), .groups = "drop") %>%   
  arrange(mean_growth_rate) %>%
  mutate(Strain = factor(Strain, levels = Strain)) %>%
  mutate(n_strains = length(unique(Strain))) 


# Percent Biofilm  
percent_biofilm_od_data_file <- update_strain_gs(percent_biofilm_od_data_file)
percent_biofilm_transformed <- percent_biofilm_od_data_file %>%
  mutate(Strain = as.character(Strain),
        arcsq_mPercentOD = asin(sqrt(mPercentOD/100))) 
#n_strains <- length(unique(percent_biofilm_transformed$Strain))
#n_observations <-length(percent_biofilm_transformed$arcsq_mPercentOD)
percent_biofilm_means <- percent_biofilm_transformed %>%
  group_by(Strain) %>%
  summarise(n_isolates = n(), mean_arcsq_mPercentOD = mean(arcsq_mPercentOD, na.rm = TRUE), 
            .groups = "drop") %>%
  arrange(mean_arcsq_mPercentOD) %>%
  mutate(Strain = factor(Strain, levels = Strain)) %>%
  mutate(n_strains = length(unique(Strain))) 

# Safranin (OD450) 
safranin_od_data_file <- update_strain_gs(safranin_od_data_file)

#n_strains <- length(unique(safranin_od_data_file$Strain))
#n_observations <-length(safranin_od_data_file$mSafOD)

safranin_means <- safranin_od_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>%
  summarise(n_isolates = n(), mean_safranin_od = mean(mSafOD, na.rm = TRUE), .groups = "drop") %>%  # adjust
  arrange(mean_safranin_od) %>%
  mutate(Strain = factor(Strain, levels = Strain)) %>%
  mutate(n_strains = length(unique(Strain))) 


# MDZ-MIC
metro_mic_data_file <- update_strain_gs(metro_mic_data_file)

metro_mic_filtered <- metro_mic_data_file %>%
  filter(!grepl("*read",
                MIC, ignore.case = TRUE)) %>%
  mutate(MIC = case_when(
    grepl("^>256$", MIC) ~ 257,  # Numeric 257, not string "257"
    TRUE ~ as.numeric(MIC)       # Convert all to numeric
  )) %>%
  filter(!is.na(MIC)) %>%
  mutate(n_strains = length(unique(Strain)))

#n_strains <- length(unique(metro_mic_filtered$Strain))
#n_observations <-length(metro_mic_filtered$MIC)

mic_means <- metro_mic_filtered %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>%
  summarise(n_isolates = n(), mean_mic = mean(MIC, na.rm = TRUE), .groups = "drop") %>%  # adjust
  arrange(mean_mic) %>%
  mutate(Strain = factor(Strain, levels = Strain))


mic_numeric <- metro_mic_filtered %>%
  mutate(
    MIC_num = case_when(
      grepl("^>256$", MIC) ~ 257,
      grepl("^[0-9]+$", MIC) ~ as.numeric(MIC),
      TRUE ~ NA_real_
    ),
    MIC_original = case_when(
      MIC_num == 257 ~ ">256",
      TRUE ~ ifelse(is.na(MIC_num), MIC, as.character(MIC_num))
    ),
    resistant = MIC_num >= 32
  ) %>%
  filter(!is.na(MIC_num))


mic_plot_data <- mic_numeric  

strain_counts <- mic_plot_data %>%
  count(Strain, sort = FALSE) %>%  
  arrange(n)  # sort least to greatest
mic_plot_data$Strain_sorted <- factor(mic_plot_data$Strain, 
                                     levels = strain_counts$Strain)


# Calculate MIC90 per strain (90th percentile of MIC values)
mic90_results <- mic_numeric %>%
  group_by(Strain) %>%
  summarise(
    n_isolates = n(),
    MIC50 = round(quantile(MIC_num, 0.5, na.rm = TRUE), 1),
    MIC90 = round(quantile(MIC_num, 0.9, na.rm = TRUE), 1),
    isolates_ge_MIC90 = sum(MIC_num >= MIC90, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(MIC90))

mic90_results


# Before/after MIC filtering
total_original <- nrow(metro_mic_data_file)
cat("Total MIC observations:", total_original, "\n")

filtered_out <- metro_mic_data_file %>%
  filter(grepl("*read",
               MIC, ignore.case = TRUE)) %>%
  count(Strain, sort = TRUE) %>%
  mutate(n_filtered = n)

total_filtered <- sum(filtered_out$n_filtered)
cat("Total MIC observations filtered out:", total_filtered, "\n")
cat("Remaining MIC observations:", total_original - total_filtered, "\n\n")
print(filtered_out)

# Create binarized dataframe from mic_numeric (MIC >=32 = 1, <32 = 0)
mic_binarized <- mic_numeric %>%
  mutate(
    MIC_binary = case_when(
      MIC_num >= 32 ~ 1,    # Resistant
      MIC_num < 32 ~ 0,     # Susceptible
      TRUE ~ NA_real_
    ),
    MIC_original = case_when(
      MIC_num == 257 ~ ">256",
      TRUE ~ as.character(MIC_num)
    )
  ) %>%
  filter(!is.na(MIC_binary))

# Verify binary distribution
table(mic_binarized$MIC_binary, useNA = "always")
prop.table(table(mic_binarized$MIC_binary))


# Mean binary resistance per strain
resistance_means <- mic_binarized %>%
  group_by(Strain) %>%
  summarise(
    prop_resistant = mean(MIC_binary),
    n_isolates = n(),
    .groups = "drop"
  ) %>%
  mutate(Strain = factor(Strain, levels = Strain[order(prop_resistant)]))
resistance_means_sorted <- resistance_means[order(resistance_means$n_isolates), ]
print(resistance_means_sorted)
```
### MDZ-MIC Threshold Reference 

Landlinger C, Oberbauer V, Podpera Tisakova L, Schwebs T, Berdaguer R, Van Simaey L, Vaneechoutte M, Corsini L. 2022. Preclinical Data on the Gardnerella-Specific Endolysin PM-477 Indicate Its Potential to Improve the Treatment of Bacterial Vaginosis through Enhanced Biofilm Removal and Avoidance of Resistance. Antimicrob Agents Chemother 66:e02319-21.
https://doi.org/10.1128/aac.02319-21

> “However, topical intravaginal antibiotic administration is recommended for BV therapy, which allows the drugs to reach concentrations of active ingredients in the mg/mL range. Therefore, in this study, we used resistance breakpoints of ≥8μg/mL and ≥32μg/mL for CLI and MDZ, respectively, as previously described for BV-related bacteria" 

Petrina MAB, Cosentino LA, Rabe LK, Hillier SL. 2017. Susceptibility of bacterial vaginosis (BV)-associated bacteria to secnidazole compared to metronidazole, tinidazole and clindamycin. Anaerobe 47:115–119. 

> "Gardnerella vaginalis (128; 64; 32)" MIC90 = 32 μg/mL 


### Using t-test to identify the minimum pH value where non-zero growth (OD600 >0.01) occurred for each strain
```{r, message=FALSE, warning=FALSE, echo=TRUE}
pH_growth_data <- pH_growth_data_file %>%
  group_by(Strain, pH) %>%
  t_test(mOD ~ 1, mu = 0.01, alternative = "greater") %>%
  add_significance("p") %>%
  filter(p.signif != "ns") %>%
  mutate(n_strains = length(unique(Strain))) %>%
  group_by(Strain) %>%
  arrange(pH) %>%
  filter(pH == min(pH)) %>%
  mutate(pH = as.numeric(pH)) 

ph_vector <- setNames(pH_growth_data$pH, pH_growth_data$Strain)
ph_range <- range(ph_vector, na.rm = TRUE)
ph_cv <- (sd(ph_vector) / mean(ph_vector))  
cat("pH_growth_data")
cat("Mean: ", mean(ph_vector), "\n")
cat("SD: ", sd(ph_vector), "\n")
cat("CV: ", ph_cv, "\n")


pH_growth_data <- update_strain_gs(pH_growth_data) 
pH_growth_data <- pH_growth_data %>%
    mutate(n_isolates = n()) 
pH_growth_data <- pH_growth_data[order(pH_growth_data$pH), ] 
#n_strains <- length(unique(pH_growth_data$Strain))
#n_observations <-length(pH_growth_data$pH)
```
# Plot replicate information for each strains (Count, Observed Values and Average per Strain)
```{r, message=FALSE, warning=FALSE, echo=TRUE}

# Carrying Capacity 
# Replicate values and averages per strain 
p1 <- ggplot(carrying_capacity_data_file, aes(x = Strain, y = maxOD)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.7) +
  geom_crossbar(data = carrying_capacity_means, aes(y = mean_maxOD, ymin = mean_maxOD, ymax = mean_maxOD),
                width = 0.5, colour = "red", size = 0.8) +
  ggtitle("Carrying Capacity", subtitle = paste0("n = ", carrying_capacity_means$n_strains[1], " strains")) +
  labs(y = "OD600") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

#Growth Rate
# Replicate values and averages per strain 
p2 <- ggplot(growth_rate_data_file, aes(x = Strain, y = mSlope)) +   
  geom_jitter(width = 0.1, height = 0, alpha = 0.7) +
  geom_crossbar(data = growth_rate_means, aes(y = mean_growth_rate, ymin = mean_growth_rate, ymax = mean_growth_rate),
                width = 0.5, colour = "red", size = 0.8) +
  ggtitle("Exponential Growth Rate", subtitle = paste0("n = ", growth_rate_means$n_strains[1], " strains")) +
  labs(y = "μ/hour") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Percent Biofilm  
p3 <- ggplot(percent_biofilm_transformed, aes(x = Strain, y = arcsq_mPercentOD)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.7) +
  geom_crossbar(data = percent_biofilm_means, 
                aes(y = mean_arcsq_mPercentOD, ymin = mean_arcsq_mPercentOD, ymax = mean_arcsq_mPercentOD),
                width = 0.5, colour = "red", size = 0.8) +
  ggtitle("Biofilm Formation", subtitle = paste0("n = ", percent_biofilm_means$n_strains[1], " strains")) +
  labs(y = "Arcsin√(% Biofilm)") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) 

# Safranin (OD450) 
p4 <- ggplot(safranin_od_data_file, aes(x = Strain, y = mSafOD)) +  
  geom_jitter(width = 0.1, height = 0, alpha = 0.7) +
  geom_crossbar(data = safranin_means, aes(y = mean_safranin_od, ymin = mean_safranin_od, ymax = mean_safranin_od),
                width = 0.5, colour = "red", size = 0.8) +
  ggtitle("Biofilm Density", subtitle = paste0("n = ", safranin_means$n_strains[1], " strains")) +
  labs(y = "Safranin OD450") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) 

# Minimum pH Growth
p5 <- ggplot(pH_growth_data, aes(x = reorder(Strain, pH), y = pH)) +  
  geom_jitter(width = 0.1, height = 0, alpha = 0.7, size = 2) +
  ggtitle("Lowest pH (3.5–6) with OD600 ≥ 0.01", subtitle = paste0("n = ", pH_growth_data$n_strains[1], " strains")) +
  labs(y = "pH") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) 

# MDZ-MIC
p6 <- ggplot(metro_mic_filtered, aes(x = Strain, y = MIC)) +
  geom_jitter(width = 0.1, height = 0, alpha = 0.7) +
  geom_crossbar(data = mic_means, aes(y = mean_mic, ymin = mean_mic, ymax = mean_mic),
                width = 0.5, colour = "red", size = 0.8) +
  ggtitle("Metronidazole Susceptibility", subtitle = paste0("n = ", metro_mic_filtered$n_strains[1], " strains")) +
  labs(y = "Minimum Inhibition Concentration (MIC)") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

p7 <- ggplot(mic_plot_data, 
                     aes(x = Strain_sorted, y = MIC_num, fill = resistant)) +
  geom_jitter(width = 0.2, height = 0, shape = 21, size = 3.5, 
              alpha = 0.7, stroke = 0.4, color = "black") +
  geom_hline(yintercept = 32, linetype = "dashed", color = "red", size = 1.2) +
  scale_fill_manual(values = c("FALSE" = "blue", "TRUE" = "red"), 
                    name = "MIC ≥32", labels = c("Susceptible", "Resistant")) +
  scale_y_log10() +
  labs(title = "Individual MIC Values by Gardnerella Strain (n=22)",
       subtitle = "Red dashed = 32 µg/mL resistance cutoff",
       x = "Strain", y = "MIC (µg/mL)") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 9, angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"))

# Plot: mean binary resistance per strain
p8 <- ggplot(resistance_means_sorted,
                             aes(reorder(Strain, n_isolates), y = prop_resistant,
                                 fill = prop_resistant > 0.5)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black", size = 1.2) +
  geom_point(shape = 21, size = 4, alpha = 1.0, stroke = 0.5) +
  scale_fill_manual(values = c("FALSE" = "blue", "TRUE" = "red"),
                    name = "Mean Binarized MIC Value", labels = c("≤0.5", ">0.5")) +
  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1)) +
  labs(title = "MDZ Susceptibility: Average of Binarized MIC", 
       subtitle = ">=32µg/mL MIC = 1 (Resistant) and <32µg/mL MIC = 0 (Susceptible), n strains = 22",
       x = "Strain", y = "Average of Binarized MIC") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  )


# Number of replicates per strain 
p9 <- ggplot(carrying_capacity_means,
             aes(x = reorder(Strain, n_isolates), y = n_isolates)) +   
  geom_col(fill = "gray", alpha = 0.8, color = "black", size = 0.2) +
  geom_text(stat = "summary", fun = "sum", 
            aes(label = after_stat(y), vjust = -0.3),
            size = 3.5, fontface = "bold", color = "black") +
    scale_y_continuous(limits = c(0, 6),
                     breaks = seq(0, 6, by = 1),
                     expand = expansion(mult = c(0, 0.15))) + 
  labs(title = "Carrying Capacity", 
       subtitle = paste0("Measured as max OD600 over incubation period; ", "n = ", carrying_capacity_means$n_strains[1], " strains"),
       x = "Strain", 
       y = "Replicate Count") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.y = element_text(size = 10)
  )


# Number of replicates per strain 
p10 <- ggplot(growth_rate_means,
             aes(x = reorder(Strain, n_isolates), y = n_isolates)) +   
  geom_col(fill = "gray", alpha = 0.8, color = "black", size = 0.2) +
  geom_text(stat = "summary", fun = "sum", 
            aes(label = after_stat(y), vjust = -0.3),
            size = 3.5, fontface = "bold", color = "black") +
    scale_y_continuous(limits = c(0, 6),
                     breaks = seq(0, 6, by = 1),
                     expand = expansion(mult = c(0, 0.15))) + 
  labs(title = "Exponential Growth Rate", 
       subtitle = paste0("n = ", growth_rate_means$n_strains[1], " strains"),
       x = "Strain", 
       y = "Replicate Count") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.y = element_text(size = 10)
  )



p11 <- ggplot(percent_biofilm_means,
             aes(x = reorder(Strain, n_isolates), y = n_isolates)) +   
  geom_col(fill = "gray", alpha = 0.8, color = "black", size = 0.2) +
  geom_text(stat = "summary", fun = "sum", 
            aes(label = after_stat(y), vjust = -0.3),
            size = 3.5, fontface = "bold", color = "black") +
    scale_y_continuous(limits = c(0, 6),
                     breaks = seq(0, 6, by = 1),
                     expand = expansion(mult = c(0, 0.15))) + 
  labs(title = "Biofilm Formation (Arcsin√(% Biofilm))", 
       subtitle = paste0("n = ", percent_biofilm_means$n_strains[1], " strains"),
       x = "Strain", 
       y = "Replicate Count") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.y = element_text(size = 10)
  )


p12 <- ggplot(safranin_means,
             aes(x = reorder(Strain, n_isolates), y = n_isolates)) +   
  geom_col(fill = "gray", alpha = 0.8, color = "black", size = 0.2) +
  geom_text(stat = "summary", fun = "sum", 
            aes(label = after_stat(y), vjust = -0.3),
            size = 3.5, fontface = "bold", color = "black") +
    scale_y_continuous(limits = c(0, 6),
                     breaks = seq(0, 6, by = 1),
                     expand = expansion(mult = c(0, 0.15))) + 
  labs(title = "Biofilm Density (Safranin OD450)", 
       subtitle = paste0("n = ", safranin_means$n_strains[1], " strains"),
       x = "Strain", 
       y = "Replicate Count") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.y = element_text(size = 10)
  )

p13 <- ggplot(resistance_means_sorted,
             aes(x = reorder(Strain, n_isolates), y = n_isolates)) +   
  geom_col(fill = "gray", alpha = 0.8, color = "black", size = 0.2) +
  geom_text(stat = "summary", fun = "sum", 
            aes(label = after_stat(y), vjust = -0.3),
            size = 3.5, fontface = "bold", color = "black") +
   scale_y_continuous(limits = c(0, 6),
                     breaks = seq(0, 6, by = 1),
                     expand = expansion(mult = c(0, 0.15))) + 
  labs(title = "Metronidazole (MDZ) Minimum Inhibition Concentration (MIC)", 
       subtitle = paste0("n = ", metro_mic_filtered$n_strains[1], " strains"),
       x = "Strain", 
       y = "Replicate Count") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.y = element_text(size = 10)
  )


p14 <- ggplot(pH_growth_data,
             aes(x = reorder(Strain, n_isolates), y = n_isolates)) +   
  geom_col(fill = "gray", alpha = 0.8, color = "black", size = 0.2) +
  geom_text(stat = "summary", fun = "sum", 
            aes(label = after_stat(y), vjust = -0.3),
            size = 3.5, fontface = "bold", color = "black") +
  scale_y_continuous(limits = c(0, 6),
                     breaks = seq(0, 6, by = 1),
                     expand = expansion(mult = c(0, 0.15))) + 
  labs(title = "Minimum pH with Growth", 
       subtitle = paste0("Lowest pH (3.5–6) with OD600 ≥ 0.01; ", "n = ", pH_growth_data$n_strains[1], " strains"),
       x = "Strain", 
       y = "Replicate Count") +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 90, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.text.y = element_text(size = 10)
  )


# View plots
p1; p2; p3; p4; p5; p6; p7; p8; p9; p10; p11; p12; p13; p14


combined_plot <- (p1 | p2 | p3) / (p4 | p6 | p5) + 
  #plot_annotation(title = "Individual and Averaged Replicate Values") +
  plot_layout(guides = "collect") &
 # theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) &  
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 8, hjust = 0.5),
    axis.text.x   = element_text(size = 8, colour = "black", face = "bold"),
    axis.text.y   = element_text(size = 8, colour = "black", face = "bold"),
    axis.title    = element_text(size = 9, face = "bold"),
    plot.margin   = margin(6, 6, 6, 6)
  )

print(combined_plot)
ggsave("strain_phenotype_averages.png", 
       plot = combined_plot, 
       width = 14, height = 10, 
       dpi = 400, 
       bg = "white")

combined_plot <- (p9 | p10 | p11) / (p12 | p13 | p14) + 
  #plot_annotation(title = "Replicate Count per Gardnerella Strain") + 
  plot_layout(guides = "collect") &
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    #plot.annotation.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 8, hjust = 0.5),
    axis.text.x   = element_text(size = 8, colour = "black", face = "bold"),
    axis.text.y   = element_text(size = 8, colour = "black", face = "bold"),
    axis.title    = element_text(size = 9, face = "bold"),
    plot.margin   = margin(6, 6, 6, 6)
  )

print(combined_plot)
ggsave("strain_phenotype_replicate_numbers.png", 
       plot = combined_plot, 
       width = 14, height = 10, 
       dpi = 400, 
       bg = "white")
```

# Average the phenotype values across replicates per strain  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
carrying_capacity_data <- carrying_capacity_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_maxOD = mean(maxOD, na.rm = TRUE))

cc_vector <- setNames(carrying_capacity_data$mean_maxOD, carrying_capacity_data$Strain)
cc_range <- range(cc_vector, na.rm = TRUE)
cc_cv_between <- (sd(cc_vector) / mean(cc_vector)) 
cat("\ncarrying_capacity_data\n")
cat("Mean: ", mean(cc_vector), "\n")
cat("SD: ", sd(cc_vector), "\n")
cat("CV: ", cc_cv_between, "\n")


growth_rate_data <- growth_rate_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_mSlope = mean(mSlope, na.rm = TRUE))
gr_vector <- setNames(growth_rate_data$mean_mSlope, growth_rate_data$Strain)
gr_range <- range(gr_vector, na.rm = TRUE)
gr_cv_between <- (sd(gr_vector) / mean(gr_vector)) 
cat("\ngrowth_rate_data\n")
cat("Mean: ", mean(gr_vector), "\n")
cat("SD: ", sd(gr_vector), "\n")
cat("CV: ", gr_cv_between, "\n")

# Transform percent_biofilm by applying asin(sqrt(x))
percent_biofilm_od_data <- percent_biofilm_od_data_file %>%
  mutate(Strain = as.character(Strain),
         arcsq_mPercentOD = asin(sqrt(mPercentOD/100))) %>%
  group_by(Strain) %>% 
  summarise(mean_arcsq_mPercentOD = mean(arcsq_mPercentOD, na.rm = TRUE))   
pb_vector <- setNames(percent_biofilm_od_data$mean_arcsq_mPercentOD, percent_biofilm_od_data$Strain)
pb_range <- range(pb_vector, na.rm = TRUE)
pb_cv_between <- (sd(pb_vector) / mean(pb_vector)) 
cat("\npercent_biofilm_od_data\n")
cat("Mean: ", mean(pb_vector), "\n")
cat("SD: ", sd(pb_vector), "\n")
cat("CV: ", pb_cv_between, "\n")

safranin_od_data <- safranin_od_data_file %>%
  mutate(Strain = as.character(Strain)) %>%
  group_by(Strain) %>% summarise(mean_mSafOD = mean(mSafOD, na.rm = TRUE))
sf_vector <- setNames(safranin_od_data$mean_mSafOD, safranin_od_data$Strain)
sf_range <- range(sf_vector, na.rm = TRUE)
sf_cv_between <- (sd(sf_vector) / mean(sf_vector)) 
cat("\nsafranin_od_data\n")
cat("Mean: ", mean(sf_vector), "\n")
cat("SD: ", sd(sf_vector), "\n")
cat("CV: ", sf_cv_between, "\n")

# MIC: Mean binary resistance per strain
mic_data <- metro_mic_data_file %>%
   filter(!grepl("*read",
                 MIC, ignore.case = TRUE)) %>%
  mutate(
    Strain = as.character(Strain),
    MIC_binary = case_when(
      grepl("^>256$", MIC) ~ 1,
      grepl("^[0-9]+$", MIC) & as.numeric(MIC) >= 32 ~ 1,
      TRUE ~ 0
    )
  ) %>%
  group_by(Strain) %>%
  summarise(
    mean_MIC_binary = mean(MIC_binary, na.rm = TRUE),
    n_isolates = n(),
    .groups = "drop"
  ) %>%
  mutate(
    Strain = factor(Strain, levels = Strain[order(mean_MIC_binary)])   
  )

mic_vector <- setNames(mic_data$mean_MIC_binary, mic_data$Strain)
mic_range <- range(mic_vector, na.rm = TRUE)

mic_cv_between <- sd(mic_vector) / mean(mic_vector)

cat("\nmic_data\n")
cat("Mean: ", mean(mic_vector), "\n")
cat("SD: ", sd(mic_vector), "\n")
cat("CV: ", mic_cv_between, "\n")
```

### Rename complete-numeric strains (e.g. 14018)  
```{r, message=FALSE, warning=FALSE, echo=TRUE}
# Strain 14018 not present in: pb_vector, sf_vector (did not pass QC for these experiments)
vecs <- list(ph=ph_vector, cc=cc_vector, gr=gr_vector, mic=mic_vector)
vecs[] <- lapply(vecs, function(v) {
  idx <- names(v) == "14018"   
  if(any(idx)) names(v)[idx] <- "S14018"   
  v
})

({cc_vector<<-vecs$cc; gr_vector<<-vecs$gr; ph_vector<<-vecs$ph; mic_vecotr<<-vecs$mic})
names(ph_vector)
```

# Phenotype Distribution Visualizations
```{r, message=FALSE, warning=FALSE}
# Okabe–Ito palette  
# Orange       #E69F00
# Sky blue     #56B4E9
# Bluish green #009E73
# Yellow       #F0E442
# Blue         #0072B2
# Vermillion   #D55E00
# Reddish purp #CC79A7
# Grey         #999999
# Black        #000000

p1 <- ggplot(carrying_capacity_data, aes(x = mean_maxOD)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "#0072B2") +  # blue
  geom_text(
    stat = "bin", binwidth = 0.1,
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    position = position_stack(vjust = 0.8),
    color = "black", size = 5
  ) +
  coord_cartesian(xlim = c(cc_range[1] - 0.05, cc_range[2] + 0.05)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.10), labels = function(x) round(x, 2)) +
  labs(
    title    = "Carrying Capacity",
   subtitle = "Measured as max OD600 over incubation period; n = 22",
    x        = "Average Max OD600",
    y        = "Number of Strains"
  ) +
  theme(
    axis.text.x   = element_text(size = 11, colour = "black"),
    axis.text.y   = element_text(size = 11, colour = "black"),
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

p2 <- ggplot(growth_rate_data, aes(x = mean_mSlope)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "#E69F00") +  # orange
  geom_text(
    stat = "bin", binwidth = 0.1,
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    position = position_stack(vjust = 0.8),
    color = "black", size = 5
  ) +
  coord_cartesian(xlim = c(gr_range[1] - 0.05, gr_range[2] + 0.05)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.10), labels = function(x) round(x, 2)) +
  labs(
    title    = "Exponential Growth Rate",
    subtitle = "n = 22",
    x        = "Average Exponential Growth Rate",
    y        = "Number of Strains"
  ) +
  theme(
    axis.text.x   = element_text(size = 11, colour = "black"),
    axis.text.y   = element_text(size = 11, colour = "black"),
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

p3 <- ggplot(pH_growth_data, aes(x = pH)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "#009E73") +  # bluish green
  geom_text(
    stat = "bin", binwidth = 0.5,
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    position = position_stack(vjust = 0.8),
    color = "black", size = 5
  ) +
  coord_cartesian(xlim = c(ph_range[1] - 0.1, ph_range[2] + 0.1)) +
  scale_x_continuous(breaks = seq(3, 6, by = 0.5), labels = function(x) round(x, 1)) +
  labs(
    title    = "Minimum pH with Growth",
    subtitle = "Lowest pH (3.5–6) with OD600 ≥ 0.01; n = 22",
    x        = "Average Minimum pH",
    y        = "Number of Strains"
  ) +
  theme(
    axis.text.x   = element_text(size = 11, colour = "black"),
    axis.text.y   = element_text(size = 11, colour = "black"),
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

p4 <- ggplot(percent_biofilm_od_data, aes(x = mean_arcsq_mPercentOD)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "#D55E00") +  # vermillion
  geom_text(
    stat = "bin", binwidth = 0.1,
    aes(label = ifelse(after_stat(count) > 0, after_stat(count), "")),
    position = position_stack(vjust = 0.8),
    color = "black", size = 4
  ) +
  coord_cartesian(xlim = c(pb_range[1] - 0.1, pb_range[2] + 0.1)) +
  scale_x_continuous(breaks = seq(0.5, 1.5, by = 0.10), labels = function(x) round(x, 2)) +
  labs(
    title    = "Biofilm Formation",
   subtitle = paste("n =", nrow(percent_biofilm_od_data)),
    x        = "Average Arcsine-Sqrt(Percent Biofilm)",
    y        = "Number of Strains"
  ) +
  theme(
    axis.text.x   = element_text(size = 11, colour = "black"),
    axis.text.y   = element_text(size = 11, colour = "black"),
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

p5 <- ggplot(safranin_od_data, aes(x = mean_mSafOD)) +
  geom_histogram(binwidth = 0.1, color = "black", fill = "#CC79A7") +  # reddish purple
  geom_text(
    stat = "bin", binwidth = 0.1,
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    position = position_stack(vjust = 0.8),
    color = "black", size = 5
  ) +
  coord_cartesian(xlim = c(sf_range[1] - 0.1, sf_range[2] + 0.1)) +
  #scale_x_continuous(breaks = pretty(sf_range, n = 10), labels = function(x) round(x, 2)) +
  scale_x_continuous(breaks = seq(0, 0.5, by = 0.10), labels = function(x) round(x, 2)) +
  labs(
    title    = "Biofilm Density",
    subtitle = "n = 21",
    x        = "Average Safranin Stain OD450",
    y        = "Number of Strains"
  ) +
  theme(
    axis.text.x   = element_text(size = 11, colour = "black"),
    axis.text.y   = element_text(size = 11, colour = "black"),
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

p6 <- ggplot(mic_data, aes(x = mean_MIC_binary)) +
  geom_histogram(
    binwidth = 0.1,
    color    = "black",
    fill     = "#F0E442",    
    boundary = 0.05
  ) +
  geom_text(
    stat     = "bin",
    binwidth = 0.1,
    aes(label = ifelse(after_stat(count) == 0, "", after_stat(count))),
    position = position_stack(vjust = 0.65),
    hjust    = 0.6,
    color    = "black",
    size     = 4.5
  ) +
  coord_cartesian(xlim = c(0, 1)) +
  scale_x_continuous(breaks = seq(0, 1, by = 0.10), labels = function(x) round(x, 2)) +
  labs(
    title    = "MDZ Susceptibility",
    subtitle = "≥32 µg/mL MIC → 1 (Resistant), <32 µg/mL MIC → 0 (Susceptible); n strains = 22",
    x        = "Average of Binarized MIC Across Strain Isolates",
    y        = "Number of Strains"
  ) +
  theme(
    axis.text.x   = element_text(size = 11, colour = "black"),
    axis.text.y   = element_text(size = 11, colour = "black"),
    plot.title    = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )

print(p1); print(p2); print(p3); print(p4); print(p5); print(p6)


```

```{r, message=FALSE, warning=FALSE}
combined_plot <- (p1 | p2 | p4) / (p5 | p6 | p3) + 
  plot_layout(guides = "collect") &
  theme(
    plot.title    = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 8, hjust = 0.5, colour = "black", face = "bold"),
    axis.text.x   = element_text(size = 8, colour = "black", face = "bold"),
    axis.text.y   = element_text(size = 8, colour = "black", face = "bold"),
    axis.title    = element_text(size = 9, face = "bold"),
    plot.margin   = margin(6, 6, 6, 6)
  )
combined_plot
print(combined_plot)
ggsave("strain_phenotype_histograms.png", 
       plot = combined_plot, 
       width = 14, height = 10, 
       dpi = 400, 
       bg = "white")
```